
---
execute:
  warning: false
  message: false
format: pdf
---

```{r}
#| label: setup
#| message: false
devtools::load_all()
library(tidyr)
library(purrr)
library(gridExtra)
library(dplyr)
library(tidyr)

# create training landscapes
#landscapes <- create_landscapes()

# Calculate metrics
#landscape_metrics <- calculate_landscape_metrics(landscapes)

# Read from disk to save time
landscape_metrics <- readr::read_csv("landscape_metrics.csv")
```

# Example of a Kruskal test with effect size

The normal Kruskal test does not contain effect size, only p-value and chisq

```{r}
# A normal kruskal test
t <- kruskal.test(
  value ~ pattern,
  data = landscape_metrics
)

t
str(t)
```

The function `rstatix::kruskal_effsize` gives us the effect size. The bigger the value for eta2 the stronger the effect. This is the value I use to rank the metrics.

```{r}
rstatix::kruskal_effsize(
  value ~ pattern,
  data = landscape_metrics
)
```


# Compare all options of metrics selections

I tried:

All methods of metrics selection:

- "coeffvar_all"
- "lin_mod_r2"
- "mean_groups"
- "fisher_score"
- "kruskal_p"
- "kruskal_effsize"

I tried them with both scaled and unscaled metrics (using `scale` in R separately for each metric).

```{r}
#| echo: false
all_options <- tidyr::expand_grid(
  selection_method = c(
    "coeffvar_all",
    "lin_mod_r2",
    "mean_groups",
    "fisher_score",
    "kruskal_p",
    "kruskal_effsize"
  ),
  scaled = c(TRUE, FALSE)
)

all_options <- all_options |>
  dplyr::rowwise() |>
  mutate(
    best_10 = pmap(
      list(selection_method, scaled),
      ~ evaluate_landscape_metrics(
        landscape_metrics,
        method = ..1,
        scale = ..2
      )
    )
  )

all_options <- all_options |>
  mutate(
    best_10_text = paste(best_10, collapse = ", ")
  )

# Make figure
all_options <- all_options |>
  mutate(
    fig_metrics = list(plot_metrics(
      metrics = landscape_metrics,
      selected_metrics = best_10,
      force = TRUE
    ))
  )

# Make figure
all_options <- all_options |>
  mutate(
    fig_metrics = list(plot_metrics(
      metrics = landscape_metrics,
      selected_metrics = best_10,
      force = TRUE
    ))
  )
```

```{r}
#| echo: false
#| tab.cap: "Best 10 metrics selected for each method and scale combination"
knitr::kable(
  all_options |>
    select(selection_method, scaled, best_10_text),
  col.names = c("Method", "Scaled?", "Best 10 selected metrics")
)
```

## Result figures to compare

```{r}
#| fig.width: 8
#| fig.height: 12
#| echo: false
for (i in unique(all_options$selection_method)) {
  # Create header text
  header_text <- sprintf(
    "Selection method: %s, scaled (top), not scaled (bottom)",
    i
  )

  # Draw plot below header
  plots_subset <- all_options |>
    filter(selection_method == i)

  combined_plot <- patchwork::wrap_plots(plots_subset$fig_metrics, ncol = 1) +
    patchwork::plot_annotation(
      title = header_text
    ) +
    patchwork::plot_layout(axis_titles = "collect")

  print(combined_plot)
}
```