---
title: "evaluate_metrics(): problems and proposed fixes"
author: "Selina Baldauf"
date: 2026-07-28
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    embed-resources: true
execute:
  cache: true
  warning: false
  message: false
---

```{r setup}
#| cache: false
# Never cache this chunk: load_all() attaches the package as a side effect, and
# side effects are not restored from cache. A cached setup chunk is skipped on
# re-render, so the package is never loaded and later chunks fail with
# "could not find function evaluate_metrics".
devtools::load_all(if (file.exists("DESCRIPTION")) "." else "..")
library(tidyverse)

all_methods <- c(
  "coeffvar_all",
  "lin_mod_r2",
  "mean_groups",
  "fisher_score",
  "kruskal_effsize"
)

# Full ranking for one method, correlation filter switched off
rank_metrics <- function(metrics, method, ...) {
  suppressWarnings(
    evaluate_metrics(
      metrics,
      method = method,
      metrics_number = n_distinct(metrics$metric),
      correlation_threshold = 1,
      ...
    )
  )
}

rank_all_methods <- function(metrics) {
  map_chr(set_names(all_methods), \(method) {
    paste(rank_metrics(metrics, method), collapse = " > ")
  })
}

# Synthetic metrics in the shape calculate_metrics() returns
make_metrics <- function(..., n_per = 40, patterns = c("A", "B", "C")) {
  tibble(
    landscape_name = sprintf("ls%03d", seq_len(n_per * length(patterns))),
    pattern = rep(patterns, each = n_per)
  ) |>
    bind_cols(as_tibble(list(...))) |>
    pivot_longer(-c(landscape_name, pattern), names_to = "metric") |>
    mutate(level = "landscape")
}

fisher_manual <- function(data) {
  overall_mean <- mean(data$value)
  groups <- data |>
    summarise(n = n(), mean_val = mean(value), sd_val = sd(value), .by = pattern)
  (sum(groups$n * (groups$mean_val - overall_mean)^2) / (nrow(groups) - 1)) /
    (sum((groups$n - 1) * groups$sd_val^2) / (sum(groups$n) - nrow(groups)))
}

plot_groups <- function(metrics) {
  ggplot(metrics, aes(x = pattern, y = value)) +
    geom_boxplot() +
    facet_wrap(~metric, scales = "free_y") +
    theme_minimal()
}

n_per <- 40
```

```{r landscapes}
set.seed(123)
metrics_real <- calculate_metrics(create_landscapes(n = 66))
```

## Constraint before anything is changed

- All paper analyses use `kruskal_effsize` → **main results are unaffected by any fix below**.
- The SI systematic tests compare **all five** methods → changing `coeffvar_all` or
  `mean_groups` means **re-running the systematic tests and updating the SI**.

## Summary

| # | Problem | Affects | Severity | Proposed action | Re-run SI? |
|---|---------|---------|----------|-----------------|-----------|
| 1 | Same criterion as `fisher_score` | `lin_mod_r2`, `fisher_score` | Medium | Document, or remove one | No |
| 2 | Breaks on metrics crossing zero | `coeffvar_all` | High | **Code fix** | Yes |
| 3 | Ignores within-group spread; explodes when mean ≈ 0 | `mean_groups` | High | **Code fix** | Yes |
| 4 | One outlier flips the ranking | `lin_mod_r2`, `fisher_score`, `mean_groups` | Medium | Document — justifies the default | No |
| 5 | Ceiling effect | `kruskal_effsize` | Medium | Document | No |
| 6 | Blind to magnitude of separation | `kruskal_effsize` | Medium | Document | No |
| 7 | Correlation pooled across patterns | all | Medium | Document | No |
| 8 | Correlated metrics added silently | all | Low | Optional code fix (attribute) | No |

---

# `lin_mod_r2` and `fisher_score`

### 1. They are the same criterion

**Problem**

- Linked by $F = \frac{R^2}{1-R^2}\cdot\frac{N-G}{G-1}$, strictly increasing in R².
- Every metric is scored on the same *N* and *G* (guaranteed by `exclude_incomplete_metrics = TRUE`) → **identical ranking**, except for ties.

**Consequence for the paper**

- "Five methods" is not accurate — there are **four distinct criteria**.
- Also explains why the SI found no strong differences between methods.

**Proposal**

- Remove one of them as they will always be the same for our case?

# `coeffvar_all`

### 2. It breaks on metrics that cross zero

**Problem**

- CV = sd/mean is only interpretable if values are positive
- clumpy has a range [-1,1]
- parfrac can become negative...
- Mean can become near 0 → CV is **huge**

**Evidence**

```{r}
set.seed(9)
landscapes_bands <- create_landscapes(
  n = 12,
  patterns = c("bands", "gaps", "spots")
)
metrics_bands <- calculate_metrics(landscapes_bands)

metrics_bands |>
  filter(metric == "pafrac") |>
  summarise(
    min = min(value, na.rm = TRUE),
    max = max(value, na.rm = TRUE),
    mean = mean(value, na.rm = TRUE),
    sd = sd(value, na.rm = TRUE),
    cv = sd / mean,
    n_na = sum(is.na(value))
  )

### pafrac is ranked first of all metrics, purely as a numerical artefact
rank_with_incomplete <- evaluate_metrics(
  metrics_bands,
  method = "coeffvar_all",
  metrics_number = 10,
  exclude_incomplete_metrics = FALSE
)

rank_with_incomplete

plot_metrics(metrics_bands, rank_with_incomplete) +
  ggtitle(
    "Metrics selected by coeffvar_all (exclude_incomplete_metrics = FALSE)"
  )
```

**Consequence for the paper**

- `exclude_incomplete_metrics = TRUE` usually saves us, because landscapemetrics returns
  `NA` for pafrac when NP < 10. It only bites when every landscape has enough patches.
- Arguably an upstream bug: a metric documented as [1, 2] returning −4.14 without warning.

**Proposal**

```{r}
rank_by_cv_fixed <- function(metrics) {
  metrics |>
    filter(!is.na(value)) |>
    summarise(
      sd_value = sd(value),
      mean_value = mean(value),
      min_value = min(value),
      .by = metric
    ) |>
    filter(min_value >= 0, mean_value > 0) |>
    mutate(cv = sd_value / mean_value) |>
    filter(is.finite(cv)) |>
    arrange(desc(cv)) |>
    pull(metric)
}
```

- Restrict the CV ranking to metrics that are actually on a ratio scale, and warn about
  the rest. Simpler and more principled than `abs(mean)`, which does not fix the
  near-zero-mean explosion.

**Check fix**

```{r}
# evaluate_metrics() drops constant metrics before ranking, so both ranking
# functions must be compared on that same preprocessed data
metrics_preprocessed <- metrics_bands |>
  filter(
    !metric %in% (metrics_bands |>
      summarise(
        sd_value = sd(value, na.rm = TRUE),
        mean_value = mean(value, na.rm = TRUE),
        .by = metric
      ) |>
      filter(is_constant_sd(sd_value, mean_value)) |>
      pull(metric))
  )

old <- rank_by_coefficient_variation(metrics_preprocessed)
new <- rank_by_cv_fixed(metrics_preprocessed)

tibble(
  n_old = length(old),
  n_new = length(new),
  only_in_old = paste(setdiff(old, new), collapse = ", "),
  only_in_new = paste(setdiff(new, old), collapse = ", ")
)
```

✅ Exactly one metric removed (`pafrac`), none added.

---

# `mean_groups`

### 3. Ignores within-group spread, and explodes near zero

**Problem**

- `rel_mean_diff = |(mean_type − mean_all) / mean_all|`, summed over patterns.
- Uses **only the means** — a metric with well-separated means but total overlap scores the same as one that separates cleanly.
- Divides by the overall mean → a metric centred on zero gets an enormous score.

**Evidence**

```{r}
make_spread <- function(seed) {
  set.seed(seed)
  make_metrics(
    tight = c(
      rnorm(n_per, 1, 0.05),
      rnorm(n_per, 2, 0.05),
      rnorm(n_per, 3, 0.05)
    ),
    medium = c(rnorm(n_per, 1, 1), rnorm(n_per, 2, 1), rnorm(n_per, 3, 1)),
    noisy = c(rnorm(n_per, 1, 8), rnorm(n_per, 2, 8), rnorm(n_per, 3, 8)),
    n_per = n_per
  )
}

metrics_spread <- make_spread(1)

ggplot(metrics_spread, aes(x = pattern, y = value)) +
  geom_boxplot() +
  facet_wrap(~metric, scales = "free_y") +
  ggtitle("Metrics with identical group means but different noise")
```

```{r}
# a single draw is unreliable, so repeat over 100 seeds and count first place (%)
first_ranked <- function(rank_fun) {
  ranked <- map_chr(1:100, \(seed) rank_fun(make_spread(seed))[1])
  ranked <- factor(ranked, levels = c("tight", "medium", "noisy"))
  as_tibble(as.list(round(100 * table(ranked) / length(ranked))))
}

map(set_names(all_methods), \(method) {
  first_ranked(\(metrics) rank_metrics(metrics, method))
}) |>
  bind_rows(.id = "method")
```

- All three metrics have the same group means (1, 2, 3); only `tight` separates them.
- `mean_groups` ranks `noisy` first most of the time — the same directional bias as
  `coeffvar_all`, just stochastic rather than deterministic.
- Cause: the score sums **absolute** deviations, so estimation noise can only inflate it
  (E|X| ≥ |E X|). The expected score is 1.0 for all three, so the noisiest metric wins.
- Reading a single seed is misleading in both directions — seed 1 puts `medium` first.

**Proposal**

```{r}
rank_by_mean_differences_fixed <- function(metrics) {
  stats_all <- metrics |>
    summarise(
      mean_all = mean(value, na.rm = TRUE),
      sd_all = sd(value, na.rm = TRUE),
      .by = metric
    )
  metrics |>
    summarise(mean_type = mean(value, na.rm = TRUE), .by = c(metric, pattern)) |>
    left_join(stats_all, by = "metric") |>
    mutate(std_diff = abs(mean_type - mean_all) / sd_all) |>
    summarise(importance_score = sum(std_diff, na.rm = TRUE), .by = metric) |>
    arrange(desc(importance_score)) |>
    pull(metric)
}
```

- Standardise by the **standard deviation** instead of the mean.
- Fixes both problems at once: accounts for spread, and never divides by ~0.
- Equivalent to summing Cohen's-d-like standardised differences.

**Check**

```{r}
first_ranked(rank_by_mean_differences_fixed)
```

On real metrics:

```{r}
reference <- rank_metrics(metrics_real, "kruskal_effsize")
agreement <- function(ranking) {
  ranking <- ranking[ranking %in% reference]
  round(cor(match(ranking, reference), seq_along(ranking), method = "spearman"), 3)
}

c(
  mean_groups_old = agreement(rank_metrics(metrics_real, "mean_groups")),
  mean_groups_fixed = agreement(rank_by_mean_differences_fixed(metrics_real)),
  fisher_score = agreement(rank_metrics(metrics_real, "fisher_score"))
)
```

✅ Both synthetic failures fixed, and agreement with `kruskal_effsize` on real
landscapes rises (higher than `fisher_score`).

---

# `kruskal_effsize` (the default)

### 4. Moment-based methods collapse under a single outlier

**Problem**

- `lin_mod_r2`, `fisher_score` and `mean_groups` are computed from means and SDs →
  sensitive to outliers, skew, heavy tails, within-group multimodality.
- `kruskal_effsize` sees only ranks → unaffected.

**Evidence**

```{r}
metrics_outlier <- metrics_spread |>
  mutate(
    value = if_else(metric == "tight" & landscape_name == "ls001", 500, value)
  )

  ggplot(metrics_outlier, aes(x = pattern, y = value)) +
  geom_boxplot() +
  facet_wrap(~metric, scales = "free_y") +
  ggtitle("Metrics with identical group means but different noise")

enframe(rank_all_methods(metrics_outlier), name = "method", value = "ranking")
```

- One value of 500 injected into the best metric demotes `tight` from **first to last**
  under `lin_mod_r2` and `fisher_score`. `kruskal_effsize` is unchanged.

**Consequence for the paper**

- This is the justification for the default of using kruskal effsize

### 5. Ceiling effect

**Problem**

- Epsilon-squared saturates. Once groups are perfectly separated, the effect size is at
  its maximum regardless of *how far apart* they are.
- Several perfectly-separating metrics therefore tie, and the order among them is
  arbitrary tie-breaking.

**Evidence**

```{r}
set.seed(10)
metrics_ceiling <- make_metrics(
  wide_gap = c(runif(n_per, 0, 1), runif(n_per, 10, 11), runif(n_per, 20, 21)),
  narrow_gap = c(runif(n_per, 0, 1), runif(n_per, 1.2, 2.2), runif(n_per, 2.4, 3.4)),
  n_per = n_per
)

metrics_ceiling |>
  summarise(
    epsilon_squared = kruskal.test(value ~ factor(pattern))$statistic /
      ((n()^2 - 1) / (n() + 1)),
    fisher = fisher_manual(pick(everything())),
    .by = metric
  )
```

- Both are perfectly separated → **identical** epsilon-squared (0.889).
- `fisher_score` still separates them: 58 300 vs 824, a 70-fold difference.

### 6. Blind to the magnitude of separation

**Problem**

- Rank-based scoring discards the size of the gap between patterns.
- A metric with a tiny but perfectly consistent separation outranks one with a huge but
  slightly overlapping separation — although the latter is more robust for a classifier
  applied to new, noisier data.

**Evidence**

```{r}
set.seed(11)
metrics_magnitude <- make_metrics(
  tiny_but_clean = c(
    rnorm(n_per, 0, 1e-4), rnorm(n_per, 5e-4, 1e-4), rnorm(n_per, 1e-3, 1e-4)
  ),
  huge_slight_overlap = c(
    rnorm(n_per, 0, 12), rnorm(n_per, 100, 12), rnorm(n_per, 200, 12)
  ),
  n_per = n_per
)

ggplot(metrics_magnitude, aes(x = pattern, y = value)) +
  geom_boxplot() +
  facet_wrap(~metric, scales = "free_y") +
  ggtitle("Metrics with identical group means but different noise")

enframe(rank_all_methods(metrics_magnitude), name = "method", value = "ranking")
```

**Proposal for 5–6**

- Document both. No code change — these are inherent to rank-based scoring and are the
  price paid for the robustness in problem 4.
- Do **not** claim "non-parametric = assumption-free". Kruskal-Wallis read as a test of
  location shift assumes similarly-shaped group distributions.

---

# Cross-cutting: the correlation filter

### 7. Correlation is pooled across all pattern types

**Problem**

- `select_metrics_correlation()` correlates metrics over **all landscapes at once**.
- Two metrics independent *within* every pattern look strongly correlated once pooled,
  because both shift with pattern → one is discarded although it carries independent
  information.

**Evidence**

```{r}
set.seed(3)
pattern_shift <- rep(c(0, 5, 10), each = n_per)
metrics_pooled <- make_metrics(
  shifted_a = pattern_shift + rnorm(3 * n_per),
  shifted_b = pattern_shift + rnorm(3 * n_per),
  independent = rnorm(3 * n_per),
  n_per = n_per
)

metrics_wide <- metrics_pooled |>
  pivot_wider(names_from = metric, values_from = value)

list(
  pooled_correlation = round(cor(metrics_wide$shifted_a, metrics_wide$shifted_b), 3),
  within_pattern = metrics_wide |>
    summarise(correlation = round(cor(shifted_a, shifted_b), 3), .by = pattern),
  selected = evaluate_metrics(metrics_pooled, metrics_number = 2)
)
```

**Proposal**

- Document as a known limitation. Pooling is arguably *correct* here — the classifier
  also sees pooled data — but it should be stated.
- Also state in the manuscript: it is the **absolute Pearson** correlation, and the test
  is **strictly greater than** the threshold.

### 8. Correlated metrics are appended to fill the requested number

**Problem**

- If fewer than `metrics_number` uncorrelated metrics exist, correlated ones are appended.
- The warning reports how many *uncorrelated* metrics were found, not how many correlated
  ones were added.
- The return value is a plain character vector with **no marking** of which are which.

**Evidence**

```{r}
set.seed(4)
base_signal <- pattern_shift + rnorm(3 * n_per)
metrics_correlated <- make_metrics(
  m1 = base_signal + rnorm(3 * n_per, 0, 0.1),
  m2 = base_signal + rnorm(3 * n_per, 0, 0.1),
  m3 = base_signal + rnorm(3 * n_per, 0, 0.1),
  m4 = base_signal + rnorm(3 * n_per, 0, 0.1),
  n_per = n_per
)

evaluate_metrics(metrics_correlated, metrics_number = 3)
```

**Proposal**

- The fill-ins are always **appended at the end**, so they are the last *k* entries —
  the manuscript sentence "the user can decide to delete these metrics" is only
  actionable if this is stated.
- Optional code fix: attach an attribute (e.g. `attr(x, "correlated_fill")`) marking them.
  Low risk, no SI impact.

---

# Manuscript changes

- "five methods" → **four distinct criteria**; note `lin_mod_r2` and `fisher_score` are the
  same criterion expressed two ways.
- Framing sentence: four rank by separability, `coeffvar_all` is a pattern-agnostic baseline.
- Drop "Assumes linear relationships" and "Assumes normally distributed data within groups";
  replace with moment-based vs rank-based **robustness**.
- Name the effect size: **epsilon-squared**, $H / \frac{n^2-1}{n+1}$.
- Correlation filter: "**absolute** correlation **greater than** 0.7".
- Justify the default with outlier robustness (problem 4), citing `pafrac` as a real case.
- State the ceiling effect and magnitude-blindness as the cost of that robustness.
- "excludes incomplete metrics automatically" → "**by default**" (it is an argument).
- Mention the second automatic exclusion: metrics with no variation across landscapes.
- Fix typos: "This functions input is receives", double space before 0.7, "Section S2.3.21".
