---
title: "Basic Workflow with ecotoneClassifyR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Basic Workflow with ecotoneClassifyR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ecotoneClassifyR)
library(tidyverse)
```

# Introduction

This vignette demonstrates the basic workflow for using the `ecotoneClassifyR` package to classify ecological ecotone patterns. The package provides tools for generating synthetic landscapes, calculating landscape metrics, training a neural network model for classification, and applying that model to new landscapes.

Ecotones are transition zones between different ecological communities. The `ecotoneClassifyR` package focuses on classifying different patterns that these transitions can take, such as sharp treelines, diffuse transitions, clustered vegetation patterns, and more.

## Basic Workflow Overview

1. Generate synthetic landscapes for training
2. Calculate landscape metrics for these landscapes
3. Evaluate and select the most informative metrics
4. Train a neural network classification model
5. Visualize model performance
6. Apply the model to classify new landscapes

Let's walk through each step.

## 1. Generating Training Landscapes

First, we'll generate a set of synthetic landscapes that represent different ecotone patterns. These will serve as our training data.

The `generate_training_landscapes()` function creates synthetic landscapes with different ecotone patterns.
The following patterns are generated:

- sharp, diffuse or curvy boundaries
- fingers
- scattered
- clustered 
- parallel sine bands

You can also control different aspects about the landscapes, such as width and height,
or which of the supported types should be included in the training landscapes.

If you want to know more about the generation of specific landscapes, check out the [Landscape Generation](landscape-generation.html) vignette.

By default, the `generate_training_landscapes()` function generates landscapes of size
100x100 pixles, including all supported landscape patterns at an equal probability.

```{r generate_landscapes}
# Generate training landscapes with different ecotone patterns
landscapes <- generate_training_landscapes(
  n = 20, # number of landscapes to generate
  seed = 42, # random seed for reproducibility
  add_rotation = TRUE # whether to add random rotations to the landscapes
)
```

You can visualize a few or all of the generated landscapes to see the different patterns:

```{r visualize_landscapes}
# Visualize a few of the generated landscapes
plot_landscape_list(landscapes)
```

## 2. Calculating Landscape Metrics

Next, we'll calculate landscape metrics for each of our training landscapes. These metrics will be used as features for the classification model.

You have the option to list all available metrics using the `list_lsm()` function from the `landscapemetrics` package:

```{r list_metrics}
# List all available landscape metrics
landscapemetrics::list_lsm()
```

Landscape metrics can be calculated at different levels: "landscape", "patch", or "class".
Here, we calculate all metrics on the landscape level. This calculation takes 
some time and the runtime increases with the number of landscapes and the number of metrics.

```{r calculate_metrics}
# Calculate landscape metrics for the generated landscapes
training_metrics <- calculate_landscape_metrics(
  landscapes,
  level = "landscape" # one of "landscape", "patch", or "class"
)

# View the first few rows
head(training_metrics)
```

## 3. Evaluating Landscape Metrics

We can now choose those metrics that are most informative for our classification task.
For this, we can use the `evaluate_metrics()` function, which provides different 
methods to evaluate and select metrics based. Here, we use 
the default method, which selects the metrics with the highest variance across all
landscapes. If you want to learn more details about the different methods, check out
the [landscape metrics vignette](landscape-metrics.html).

```{r evaluate_metrics}
# Find the 10 best metrics
best_10 <- evaluate_landscape_metrics(
  training_metrics,
  metrics_number = 10, # number of metrics to select
  method = "coeffvar_all" # one of: "coeffvar_all", "lin_mod_r2", "mean_groups"
)
# Display the selected metrics
best_10
```

### Caveat

Right now, this function is not perfect. We aim for metrics that are informative,
but that also do not contain missing values. If we have a missing value in just one metric, the landscape will be excluded from model training. Until this is fixed, 
I recommend using all metrics but removing the ones that either contain NAs or 
are constant across all landscapes.

```{r remove_constant_metrics}
# Find metrics with NA values
na_metrics <- training_metrics |>
  filter(is.na(value) | is.nan(value)) |>
  pull(metric) |>
  unique()

# Remove metrics with NA values
training_metrics_filtered <- training_metrics |>
  filter(!(metric %in% na_metrics))

# Find metrics that are constant (zero standard deviation)
constant_metrics <- training_metrics_filtered |>
  summarise(sd = sd(value, na.rm = TRUE), .by = metric) |>
  filter(sd == 0) |>
  pull(metric)

# Remove constant metrics
training_metrics_filtered <- training_metrics_filtered |>
  filter(!(metric %in% constant_metrics))

# See how many metrics we have left
length(unique(training_metrics_filtered$metric))
```

## 4. Training the Neural Network Model

Now we'll train a neural network model using our filtered metrics.

The `train_nn_metrics()` function trains a neural network model with cross-validation.
We can optionally specify the metrics to use with the `metrics_selected` argument.
Also, we can specify the cross validation method (kfold or leave-one-out) and the number of folds.
For a low number of data points, the function will automatically use leave-one-out cross-validation.

```{r train_model}
# Train neural network model with cross-validation
model <- train_nn_metrics(
  metrics = training_metrics_filtered,
  cv_folds = 5,
  cv_method = "k-fold" # one of "k-fold", "loo", "none"
)
```

The trained model comes as an object including cross-validation results and performance checks. Just have a look at the model object:

```{r model_summary}
# Display model summary
model
```

## 5. Visualizing Model Performance

We can visualize the performance of our model using several different plots:

```{r visualize_performance}
# Or get all plots at once
all_plots <- plot_classification_results(model, return_all = TRUE)
patchwork::wrap_plots(all_plots)

# Or plot individual plots
# Plot confusion matrix
# plot_classification_results(model, plot_type = "confusion")

# Plot class probabilities
# plot_classification_results(model, plot_type = "probabilities")

# Plot classification confidence
# plot_classification_results(model, plot_type = "confidence")

# Plot misclassifications
# plot_classification_results(model, plot_type = "misclassifications")
```

## 6. Applying the Model to New Landscapes

Finally, we can apply our trained model to classify new landscapes. For this, we first need to generate new landscapes and calculate their metrics.

```{r apply_model}
# Generate a new test landscape
test_landscape <- create_landscape(
  pattern = "clustered",
  width = 100,
  height = 100,
  treeline_position = 0.5,
  num_clusters = 10,
  cluster_radius = 5,
  seed = 43,
  rotation = 0,
  add_metadata = TRUE
)

# Visualize the test landscape
plot_landscape(test_landscape)

# Classify the landscape
classification <- apply_nn_metrics(
  landscape = test_landscape,
  nn_model = model
)

# Display classification results
classification
```

We can also apply our model to a list of landscapes:

```{r apply_model_list}
# Generate multiple test landscapes
test_landscapes <- generate_training_landscapes(
  n = 5,
  seed = 43,
  add_rotation = TRUE
)

# Visualize test landscapes
plot_landscape_list(test_landscapes)

# Classify all landscapes
classifications <- apply_nn_metrics(
  landscape = test_landscapes,
  nn_model = model
)

# Display classification results
classifications
```



