---
title: "classify-landscapes-keras"
vignette: >
  %\VignetteIndexEntry{classify-landscapes-keras}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

```{r}
#| label: setup
library(spatPatClassifyR)
set.seed(123456)
```

This vignette demonstrates how to classify landscape patterns using a convolutional neural network (Keras backend) trained directly on landscape raster data.

The workflow consists of:

1. Generating training landscapes with known patterns using the `create_training_landscapes()` function.
2. Training a neural network on the landscape data using the `train_nn_landscapes()` function.
3. Classify new landscapes using the trained model with the `apply_nn_landscapes()` function.

## Setup Keras

Models are trained using the R package `keras3`. It requires a working installation of TensorFlow.

:::{.callout-warning}

This is still missing and I need to add it

:::

## Step 1: Generate Training Landscapes


```{r}
#| label: generate-landscapes
training_landscapes <- create_training_landscapes(
  n = 100,
  patterns = c("sharp", "diffuse", "clustered", "fingers", "bands", "random")
)
```

## Step 2: Train Neural Network Classifier

The model can be trained with or without cross-validation. Here, we use 5-fold cross-validation to evaluate model performance during training.
If cross-validation is used, performance will be returned and printed to the console. For details and more options, see the documentation of `train_nn_landscapes()`.

```{r}
#| label: train-model
model <- train_nn_landscapes(
  landscapes = training_landscapes,
  cv_method = "k-fold",
  cv_folds = 3
)
```

## Step 3: Apply Model to New Landscapes

```{r}
#| label: test-landscapes
test_landscapes <- create_training_landscapes(
  n = 20,
  patterns = c("sharp", "diffuse", "clustered", "fingers", "bands", "random"),
  add_rotation = TRUE
)

validation_results <- apply_nn_landscapes(
  landscapes = test_landscapes,
  nn_model = model
)
```

We can plot the classified landscapes along with their true and predicted patterns.

```{r}
#| label: plot-results
plot_classified_landscapes(
  classification = validation_results$predictions,
  landscapes = test_landscapes,
  only_misclassified = FALSE
)
```

