---
title: "Classify landscapes using Keras on landscape rasters"
vignette: >
  %\VignetteIndexEntry{classify-landscapes-keras}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

This vignette shows how to classify landscape patterns using a convolutional neural network (Keras backend) trained directly on landscape raster data.

```{r}
#| label: setup
library(spatPatClassifyR)
```

The workflow consists of the following steps:

1. Generate training landscapes with known patterns using the `create_landscapes()` function.
2. Train a neural network on the landscape pixel data using the `train_nn_pixels()` function.
3. Classify new landscapes using the trained model with the `apply_nn_pixels()` function.

## Setup Keras

Models are trained using the R package `keras3`. It requires a working installation of TensorFlow.

You can refer to our [installation guide](install-keras.html) and the official [`keras3` package website](https://keras3.posit.co/) for instructions on how to install Keras and TensorFlow.

To quickly test if your setup is working, you can run the following simple function from the `keras3` package:

```{r}
#| label: test-keras
keras3::to_categorical(0)
```

If the function does not return an error, your Keras installation is working correctly.

## Set seed for reproducibility

Keras models do not use R's random number generator (RNG). Therefore, it is not enough to set a seed with `set.seed()`.
To set both a seed for R and for Keras RNG, you can use the `set_random_seed()` function from `spatPatClassifyR`.

```{r}
#| label: set-seed
set_random_seed(123456)
```

## Step 1: Generate Training Landscapes

You can generate a set of training landscapes with known patterns.
See [landscape generation vignette](landscape-generation.html) for details on landscape generation and available patterns and options.

```{r}
#| label: generate-landscapes
training_landscapes <- create_landscapes(
  n = 100,
  patterns = c("labyrinth", "random", "clustered")
)
```

## Step 2: Train Neural Network

The model can be trained with or without cross-validation. Valid methods for 
the cross-validation method `cv_method` are:

- `"none"`: No cross-validation, train on all data.
- `"kfold"`: k-fold cross-validation with `cv_folds` number of folds.
- `"loo"`: leave-one-out cross-validation.

For details and further options see the function help of `train_nn_pixels()`.

:::{callout-warning}

Description of model architecture still missing.

:::

Here, we train the model using 2-fold cross-validation to keep computational time
low.

:::{.callout-note}
## Low fold accuaracy

The function will print fold accuracies during training (see below). If the accuracy is low,
consider increasing the number of training landscapes or check the function help for other options to improve model performance.
:::

```{r}
#| label: train-model
#| message: false
#| output: false
model <- train_nn_pixels(
  landscapes = training_landscapes,
  cv_method = "k-fold",
  cv_folds = 2,
  verbose = FALSE
)
```

To check the model performance, we can look at the confusion matrix from cross-validation:

```{r}
#| label: view-confusion-matrix
# Confusion matrix from cross-validation
model$performance$confusion_matrix
```

You can also check other performance metrics like overall accuracy:

```{r}
#| label: view-overall-accuracy
# Overall accuracy
model$performance$accuracy
```

Or per class metrics like precision, recall, and F1-score:

```{r}
#| label: view-per-class-metrics
model$performance$per_class_metrics
```

## Step 3: Classify New Landscapes

Finally, you can create some new test landscapes and classify them using the trained model. In this example, we create 20 new landscapes for testing. 

In reality this could be landscapes read in from files or created in other ways.
For details on importing own landscapes, see the [importing landscapes vignette](importing-landscapes.html).

```{r}
#| label: test-landscapes
test_landscapes <- create_landscapes(
  n = 20,
  patterns = c("labyrinth", "random", "clustered")
)
```

If test landscapes have been created with `create_landscapes()`, their true patterns are known. Therefore they can be used to evaluate classification performance.

:::{.callout-note}
To get additional performance metrics, set `return_performance = TRUE` when applying the model. This only works if true patterns are known. If true patterns are not known, set `return_performance = FALSE` (default).
:::


```{r}
#| label: apply-model
#| output: false
classification <- apply_nn_pixels(
  landscapes = test_landscapes,
  nn_model = model,
  return_performance = TRUE
)
```

You can look at the predicted patterns for each test landscape individually and
compare actual and predicted classes:

```{r}
#| label: view-predictions-2
# Predicted patterns
classification$predictions
```

You can also look at performance summaries like confusion matrix:

```{r}
#| label: view-performance-2
# Performance summary
classification$performance$confusion_matrix
```

And other metrics like accuracy, precision, recall, and F1-score:

```{r}
#| label: view-per-class-metrics-2
# Other performance metrics
classification$performance$per_class_metrics
```

To visualize the classified landscapes along with their true and predicted patterns
use the function `plot_classified_landscapes`. Correctly classified landscapes are shown in green, misclassified ones in red. To plot only misclassified landscapes, set `only_misclassified = TRUE`.

:::{.callout-note}
If you classified landscapes where the true patterns are not known, the plot will
show the landscape and the predicted pattern only.
:::

```{r}
#| label: plot-classified-landscapes
#| fig.width: 8
#| fig.height: 8
# Visualize true and predicted patterns

plot_classified_landscapes(
  classification = classification$predictions,
  landscapes = test_landscapes
)
```
