---
title: "classify-metrics"
vignette: >
  %\VignetteIndexEntry{classify-metrics}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

```{r}
#| label: setup
library(spatPatClassifyR)

# Set seed for reproducibility
set.seed(123456)
```

This vignette provides an overview of the workflow that can be used to train a neural network on landscape metrics and use the trained network to classify landscapes.

The workflow consists of the following steps:

1. Generate training landscapes with known patterns using the `create_training_landscapes()` function.
2. Calculate landscape metrics for the training landscapes using the `calculate_landscape_metrics()` function.
3. Select most informative metrics using the `evaluate_landscape_metrics()` function.
4. Train a neural network using the selected metrics with the `train_nn_metrics()` function.
5. Classify new landscapes using the trained neural network with the `apply_nn_metrics()` function.


## Step 1: Generate Training Landscapes

Here we generate a set of training landscapes with known patterns.


```{r}
#| label: generate-landscapes
# Generate 100 training landscapes
landscapes <- create_training_landscapes(
  n = 100,
  patterns = c("labyrinth", "random", "clustered"),
  add_rotation = TRUE
)
```

## Step 2: Calculate Landscape Metrics

Next, we calculate landscape metrics for the generated training landscapes. Here we use class-level metrics, but landscape-level metrics can also be calculated by setting `level = "landscape"`.


```{r}
#| label: calculate-metrics
# Calculate landscape metrics on the landscape level
landscape_metrics <- calculate_landscape_metrics(
  landscapes,
  level = "class"
)
```

## Step 3: Evaluate Landscape Metrics

Now we select the 10 most informative metrics for classification based on their relative variablility across landscapes (for more details see `vignette("landscape-metrics")`). We can plot the differences between patterns for the selected metrics using `plot_metrics`

```{r}
#| label: evaluate-metrics
best_10 <- evaluate_landscape_metrics(
  metrics = landscape_metrics,
  method = "kruskal_p",
  metrics_number = 10,
  verbose = FALSE
)

# Plot selected metrics
plot_metrics(
  metrics = landscape_metrics,
  selected_metrics = best_10
)
```

## Step 4: Train Neural Network

The model can be trained using the selected metrics.
Training can be done with or without cross-validation.
Valid options for `cv_method` are `"none"`, `"k-fold"` (k-fold cross-validation) and `"loo"` (leave-one-out cross-validation).
For details and further options see the function help `?train_nn_metrics()`.


```{r}
#| label: train-model
# Train neural network with k-fold cross-validation (3 folds)
model <- train_nn_metrics(
  metrics = landscape_metrics,
  metrics_selected = best_10,
  cv_method = "k-fold",
  cv_folds = 3,
  verbose = FALSE
)
```

To check the model performance, we can look at the confusion matrix from cross-validation:

```{r}
# Confusion matrix from cross-validation
model$performance$confusion_matrix
```

To access other performance metrics, see `model$performance`.

## Step 5: Classify New Landscapes

Finally, we can create some new test landscapes and classify them using the trained model. In this example, we create 10 new landscapes for testing. In reality this could be landscapes read in from files or created in other ways.

```{r}
#| label: classify-landscapes
# Create 10 new test landscapes
test_landscapes <- create_training_landscapes(
  n = 10,
  patterns = c("labyrinth", "random", "clustered")
)
# Classify test landscapes using the trained model
predictions <- apply_nn_metrics(test_landscapes, model)
```

We can look at the predicted patterns for the test landscapes:

```{r}
# Predicted patterns
predictions
```

We can also visualize the classified landscapes along with their true and predicted patterns. Correctly classified landscapes are shown in green, misclassified ones in red. To plot only missclassified landscapes, set `only_misclassified = TRUE`.

```{r}
# Visualize true and predicted patterns
plot_classified_landscapes(
  classification = predictions,
  landscapes = test_landscapes
)
```