---
title: "Classify landscapes using landscape metrics"
vignette: >
  %\VignetteIndexEntry{classify-metrics}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

This vignette provides an overview of the workflow to train a neural network on landscape metrics and then use the trained network to classify new landscapes based on the same metrics.

```{r}
#| label: setup
library(spatPatClassifyR)

# Set seed for reproducibility
set.seed(123456)
```


The workflow consists of the following steps:

1. Generate training landscapes with known patterns using the `create_landscapes()` function.
2. Calculate landscape metrics for the training landscapes using the `calculate_landscape_metrics()` function.
3. Select most informative metrics (i.e. metrics that differ the most between landscapes) using the `evaluate_landscape_metrics()` function.
4. Train a neural network using the selected metrics with the `train_nn_metrics()` function.
5. Classify new landscapes using the trained neural network with the `apply_nn_metrics()` function.


## Step 1: Generate Training Landscapes

You can generate a set of training landscapes with known patterns.
See [landscape generation vignette](landscape-generation.html) for details on landscape generation and available patterns and options.

```{r}
#| label: generate-landscapes
# Generate 100 training landscapes of 3 patterns
landscapes <- create_landscapes(
  n = 100,
  patterns = c("labyrinth", "random", "clustered")
)
```

## Step 2: Calculate Landscape Metrics

Next, you can calculate landscape metrics for the training landscapes. 
Landscape metrics are calculated using the metrics and functions from the [`landscapemetrics`](https://r-spatialecology.github.io/landscapemetrics/) package.
They can be calculated on the class level (i.e. for each land cover class separately) or on the landscape level (i.e. for the whole landscape).
For details, see [landscape metrics vignette](landscape-metrics.html).

In this example, we calculate metrics on the landscape level.


```{r}
#| label: calculate-metrics
#| message: false
# Calculate landscape metrics on the landscape level
landscape_metrics <- calculate_landscape_metrics(
  landscapes,
  level = "landscape"
)
```

## Step 3: Evaluate Landscape Metrics

You can use the `evaluate_landscape_metrics` function to select the n (e.g. n=10) most informative metrics for classification based on different methods. Here, we use the effect size of a Kruskal-Wallis H test (for more details and other options see [landscape metrics vignette](landscape-metrics.html)).

:::{.callout-note}
Note, that some metrics may return `NA` for some landscapes. Also, some metrics may
have the same value for each landscape (zero variance between the values). By default, these
metrics are removed before evaluation and the user gets warned about the 
metrics that are removed.

It is not recommended to set `exclude_na = FALSE`, as metrics with `NA` values cannot be used for training the neural network. They may however be informative and useful for other types of analyses.
:::


```{r}
#| label: evaluate-metrics
best_10 <- evaluate_landscape_metrics(
  metrics = landscape_metrics,
  method = "kruskal_effsize",
  metrics_number = 10,
  verbose = FALSE
)
```

Alternatively, you can manually select the metrics you want to use for further evaluation. For this, just provide a list of the metric names you are interested in. 

You can plot the differences between patterns for the selected metrics using the `plot_metrics` function.

```{r}
#| label: plot-metrics
#| fig.width: 8
#| fig.height: 6
#| fig.cap: "Boxplots show the distribution of the selected landscape metrics across different spatial patterns. Each point represents one training landscape and each panel is one of the selected metrics."
# Plot selected metrics
plot_metrics(
  metrics = landscape_metrics,
  selected_metrics = best_10
)
```

## Step 4: Train Neural Network

The model can now be trained using the selected metrics from step 3.
Training can be done with or without cross-validation.
Valid options for the cross-validation method `cv_method` are:

- `"none"`: no cross-validation
- `"k-fold"`: k-fold cross-validation with `cv_folds` number of folds
- `"loo"`: leave-one-out cross-validation

For details and further options see the function help `?train_nn_metrics()`.

```{r}
#| label: train-model
# Train neural network with k-fold cross-validation (3 folds)
model <- train_nn_metrics(
  metrics = landscape_metrics,
  metrics_selected = best_10,
  cv_method = "k-fold",
  cv_folds = 3,
  verbose = FALSE
)
```

To check the model performance, you can look at the confusion matrix from cross-validation:

```{r}
#| label: view-confusion-matrix
# Confusion matrix from cross-validation
model$performance$confusion_matrix
```

You can also check other performance metrics like overall accuracy:

```{r}
#| label: view-overall-accuracy
# Overall accuracy
model$performance$accuracy
```

Or per class metrics like precision, recall, and F1-score:

```{r}
#| label: view-per-class-metrics
model$performance$per_class_metrics
```

## Step 5: Classify New Landscapes

Finally, you can create some new test landscapes and classify them using the trained model. In this example, we create 10 new landscapes for testing. 

In reality these could be landscapes read in from files or created in other ways (e.g. outputs of simulation models).
For details on importing own landscapes, see the [importing landscapes vignette](importing-landscapes.html).

```{r}
#| label: classify-landscapes
# Create 10 new test landscapes using the same patterns as for training
test_landscapes <- create_landscapes(
  n = 10,
  patterns = c("labyrinth", "random", "clustered")
)
```

If test landscapes have been created with `create_landscapes()`, their true patterns are known. Therefore they can be used to evaluate classification performance.

:::{.callout-note}
To get additional performance metrics, set `return_performance = TRUE` when applying the model. This only works if true patterns are known. If true patterns are not known, set `return_performance = FALSE` (default).
:::

```{r}
#| label: apply-model
#| message: false
#| warning: false
# Classify test landscapes using the trained model
classification <- apply_nn_metrics(
  landscapes = test_landscapes,
  nn_model = model,
  return_performance = TRUE
)
```

You can look at the predicted patterns for each test landscape individually and
compare actual and predicted classes:

```{r}
#| label: view-predictions-2
# Predicted patterns
classification$predictions
```

You can also look at performance summaries like confusion matrix:

```{r}
#| label: view-performance-2
# Performance summary
classification$performance$confusion_matrix
```

And other metrics like accuracy, precision, recall, and F1-score:

```{r}
#| label: view-per-class-metrics-2
# Other performance metrics
classification$performance$per_class_metrics
```

To visualize the classified landscapes along with their true and predicted patterns
use the function `plot_classified_landscapes`. Correctly classified landscapes are shown in green, misclassified ones in red. To plot only misclassified landscapes, set `only_misclassified = TRUE`.

:::{.callout-note}
If you classified landscapes where the true patterns are not known, the plot will
show the landscape and the predicted pattern only.
:::

```{r}
#| label: plot-classified-landscapes
#| fig.width: 8
#| fig.height: 6
# Visualize true and predicted patterns

plot_classified_landscapes(
  classification = classification$predictions,
  landscapes = test_landscapes
)
```