---
title: "Classify landscapes using Keras on landscape rasters"
vignette: >
  %\VignetteIndexEntry{classify-landscapes-keras}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
    cache: true
---

This vignette shows how to classify landscape patterns using a convolutional neural network (Keras backend) trained directly on landscape raster data.

```{r}
#| label: setup
library(spatPatClassifyR)
```

The workflow consists of the following steps:

1. Generate training landscapes with known patterns using the `create_landscapes()` function.
2. Train a neural network on the landscape pixel data using the `train_pixel_model()` function.
3. Classify new landscapes using the trained model with the `apply_pixel_model()` function.

## Setup Keras

Models are trained using the R package `keras3`. It requires a working installation of TensorFlow.

You can refer to our [installation guide](install-keras.html) and the official [`keras3` package website](https://keras3.posit.co/) for instructions on how to install Keras and TensorFlow.

Use the package's backend check to confirm that `keras3` can reach TensorFlow:

```{r}
#| label: test-keras
keras_available()
```

This must return `TRUE` before the pixel workflow can run.

## Set seed for reproducibility

The pixel workflow uses R's random-number generator for operations such as
cross-validation fold assignment and data shuffling. Keras and TensorFlow use
their own randomness for weight initialization and dropout. Use `set.seed()` to
reproduce landscape generation, and call `set_random_seed()` immediately before
training to reset both R and Keras.

```{r}
#| label: set-seed
set.seed(123456)
```

## Step 1: Generate Training Landscapes

You can generate a set of training landscapes with known patterns.
See [landscape generation vignette](landscape-generation.html) for details on landscape generation and available patterns and options.

```{r}
#| label: generate-landscapes
training_landscapes <- create_landscapes(
  n = 100,
  patterns = c("labyrinth", "random", "clustered")
)
```

## Step 2: Train Neural Network

The model can be trained with or without cross-validation. Valid methods for 
the cross-validation method `cv_method` are:

- `"none"`: No cross-validation, train on all data.
- `"k-fold"`: k-fold cross-validation with `cv_folds` number of folds.
- `"loo"`: leave-one-out cross-validation.

For details and further options see the function help of `train_pixel_model()`.

The only built-in architecture, `"multiscale"`, processes the landscapes as raster
arrays. It has two sequential convolutional blocks. Each block applies 3 by 3 and
5 by 5 convolutions followed by max pooling, with 32 filters in the first block and
64 in the second. Dropout and a dense layer precede the softmax output. Users can
tune the learning rate, epochs, batch size, dropout, dense-layer size, and training
validation settings; see `?train_pixel_model` for the complete interface.

Here, we train the model using 2-fold cross-validation to keep computational time
low.

:::{.callout-note}
## Low fold accuracy

The function will print fold accuracies during training (see below). If the accuracy is low,
consider increasing the number of training landscapes or check the function help for other options to improve model performance.
:::

```{r}
#| label: train-model
#| message: false
#| output: false
set_random_seed(123456)
model <- train_pixel_model(
  landscapes = training_landscapes,
  cv_method = "k-fold",
  cv_folds = 2,
  verbose = FALSE
)
```

To check the model performance, we can look at the confusion matrix from cross-validation:

```{r}
#| label: view-confusion-matrix
# Confusion matrix from cross-validation
model$performance$confusion_matrix
```

You can also check other performance metrics like overall accuracy:

```{r}
#| label: view-overall-accuracy
# Overall accuracy
model$performance$accuracy
```

Or per class metrics like precision, recall, and F1-score:

```{r}
#| label: view-per-class-metrics
model$performance$per_class_metrics
```

## Step 3: Classify New Landscapes

Finally, you can create some new test landscapes and classify them using the trained model. In this example, we create 20 new landscapes for testing. 

In reality this could be landscapes read in from files or created in other ways.
For details on importing own landscapes, see the [importing landscapes vignette](importing-landscapes.html).

Application landscapes must contain categorical values, have no `NA` cells, and
have the same number of layers as the training data. `apply_pixel_model()`
resamples their rows and columns to the training dimensions with nearest-neighbor
resampling. It warns when a different aspect ratio means that this stretches the
pattern. See [Matching training and application data](matching-training-application.html)
for the practical consequences of resizing.

```{r}
#| label: test-landscapes
test_landscapes <- create_landscapes(
  n = 20,
  patterns = c("labyrinth", "random", "clustered")
)
```

If test landscapes have been created with `create_landscapes()`, their true patterns are known. Therefore they can be used to evaluate classification performance.

The result always has the same two parts, `$predictions` and `$performance`.

:::{.callout-note}
Performance is evaluated automatically when the true patterns are known. Landscapes that carry a `pattern` but you don't want to evaluate model performance with them but only classify them can skip evaluation with `evaluate = "none"`. Setting `evaluate = "required"` raises an error instead of quietly returning no performance when the true patterns turn out to be missing in some or all cases.
:::


```{r}
#| label: apply-model
#| output: false
classification <- apply_pixel_model(
  landscapes = test_landscapes,
  nn_model = model
)
```

You can look at the predicted patterns for each test landscape individually and
compare actual and predicted classes:

```{r}
#| label: view-predictions-2
# Predicted patterns
classification$predictions
```

The `predictions` table contains the predicted class, its `score`, and one score
for every trained class. The class scores are non-negative and sum to one, but
they are not calibrated probabilities. A score of 0.8 means that the model
assigned 80% of its total output for this landscape to the predicted class and
20% to all other classes combined. It does not mean that the prediction has an
80% probability of being correct. Use the scores to judge how decisive the model
was. A dominant score indicates stronger support for one class, while similar
scores for several classes indicate that the landscape was ambiguous to the
model.

You can also look at performance summaries like confusion matrix:

```{r}
#| label: view-performance-2
# Performance summary
classification$performance$confusion_matrix
```

And other metrics like accuracy, precision, recall, and F1-score:

```{r}
#| label: view-per-class-metrics-2
# Other performance metrics
classification$performance$per_class_metrics
```

To visualize the classified landscapes along with their true and predicted patterns
use the function `plot_classified_landscapes`. Correctly classified landscapes are shown in blue, misclassified ones in bold orange. To plot only misclassified landscapes, set `only_misclassified = TRUE`.

:::{.callout-note}
If you classified landscapes where the true patterns are not known, the plot will
show the landscape and the predicted pattern only.
:::

```{r}
#| label: plot-classified-landscapes
#| fig-width: 8
#| fig-height: 8
# Visualize true and predicted patterns

plot_classified_landscapes(
  classification = classification$predictions,
  landscapes = test_landscapes
)
```
