---
title: "Classify landscapes with landscape metrics"
vignette: >
  %\VignetteIndexEntry{Classify landscapes with landscape metrics}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
    cache: true
---

This guide shows how to train a classifier on landscape metrics and apply it to
previously unseen landscapes. The selected metrics describe landscape
composition and spatial configuration, so they can also provide insight into
which landscape characteristics distinguish the pattern classes. The workflow
uses a feed-forward neural network from the `neuralnet` package.

The example extends the metric-based quick example in
[Get started with patternscaper](patternscaper.html) by distinguishing three
ecotone patterns: sharp, clustered, and bands.

```{r}
#| label: setup
library(patternscaper)

# Set seed for reproducible example
set.seed(123456)
```

## Overview

The metric-based workflow has five steps:

1. Prepare labelled training landscapes with known patterns
2. Calculate their landscape metrics with `calculate_metrics()`
3. Select informative metrics with `evaluate_metrics()`
4. Train the final classifier with `train_metric_model()`
5. Classify new landscapes with `apply_metric_model()` and evaluate the predictions when their true patterns are known


## Step 1: Create training landscapes

Create training landscapes with known patterns. The
[artificial landscape guide](landscape-generation.html) explains how to create
training batches, and the [pattern gallery](pattern-gallery.html) shows the
available patterns and their parameters. Alternatively, import your own raster
or matrix data as described in [Import user-defined landscapes](importing-landscapes.html).

```{r}
#| label: generate-landscapes
# Generate 100 training landscapes from 3 ecotone patterns
training_landscapes <- create_landscapes(
  n = 100,
  patterns = c("sharp", "clustered", "bands")
)
```

Training landscapes should represent the pattern classes and the variation
expected in the application data.

## Step 2: Calculate landscape metrics

Next, calculate landscape metrics for the training landscapes. The `calculate_metrics()` function internally uses the functions provided by the
[`landscapemetrics`](https://r-spatialecology.github.io/landscapemetrics/)
package. Metrics can be calculated at the landscape or at the class level. At the landscape level, each metric summarizes the complete landscape, while at the class level, each metric is calculated separately for every land-cover
class.

Landscape-level metrics are suitable for the three ecotone classes in this
example. However, if patterns differ mainly by exchanging land-cover values, such as spots
and gaps or bare and dense landscapes, class-level metrics are more suitable because
landscape-level metrics may not be able to distinguish them. For more details, see
[Calculate and evaluate landscape metrics](landscape-metrics.html).


```{r}
#| label: calculate-metrics
#| message: false
# Calculate landscape metrics on the landscape level
landscape_metrics <- calculate_metrics(
  training_landscapes,
  level = "landscape"
)
```

## Step 3: Evaluate landscape metrics

Use `evaluate_metrics()` to rank metrics by how well they distinguish the
training patterns and select a number of informative metrics. This example uses the default
Kruskal-Wallis effect-size method to get the ten most informative metrics. The function reduces redundancy by filtering out highly correlated metrics. If too few metrics
pass the correlation filter, the highest-ranked correlated candidates are added
to reach the requested number of metrics. See
[Calculate and evaluate landscape metrics](landscape-metrics.html) for the
other ranking methods and details of the selection process.

:::{.callout-note}
## Incomplete metrics

By default, `evaluate_metrics()` excludes incomplete metrics that are not
available for every landscape and zero-variation metrics that have the same
value across landscapes. Incomplete metrics cannot be passed directly to the
neural network because every predictor needs a value for every training
landscape. Zero-variation metrics provide no information for distinguishing
pattern types. Leave `exclude_incomplete_metrics = TRUE` for model training.
Excluded metrics are still recorded and reported to the user but are not
selected for training.
:::


```{r}
#| label: evaluate-metrics
metric_selection <- evaluate_metrics(
  metrics = landscape_metrics,
  method = "kruskal_effsize",
  metrics_number = 10,
  verbose = FALSE
)

metric_selection
```

`evaluate_metrics()` returns the selected metric names in `$selected` together with the
ranking information in `$ranking`, which records the score of every metric and what
happened to the ones that were not selected. See the
[landscape metrics vignette](landscape-metrics.html) for how to read it. The
functions below accept either the object `metric_selection` itself or the names of the selected metrics `metric_selection$selected`.

Alternatively, provide a character vector of metric names to select them
manually.

Use `plot_metrics()` to compare the selected metric distributions among the
pattern classes.

```{r}
#| label: plot-metrics
#| fig-width: 8
#| fig-height: 6
#| fig-alt: "Ten panels of boxplots and points comparing selected landscape metrics among sharp, clustered, and bands ecotone classes"
#| fig-cap: "Distributions of selected metrics across sharp, clustered, and bands ecotone classes. Each point is one training landscape."
plot_metrics(
  metrics = landscape_metrics,
  selected_metrics = metric_selection,
  metric_labels = "name"
)
```

## Step 4: Train the final model

`train_metric_model()` fits a feed-forward neural network using the selected
metrics from step 3 as predictors. Before training, it centers and scales the
metrics because their units and ranges can differ widely. The model stores these
transformations so they can be applied to new landscapes in the same way.

Here, the final model is trained on all training landscapes without
cross-validation. Its performance will be evaluated on independent test
landscapes in step 5. Make sure to reset R's random-number generator immediately before
training so the result does not depend on random draws made by earlier steps.

See `?train_metric_model` for architecture and training options.

```{r}
#| label: train-model
# Reset the seed immediately before training
set.seed(123456)

# Train the final model on all training landscapes
model <- train_metric_model(
  metrics = landscape_metrics,
  metrics_selected = metric_selection,
  cv_method = "none",
  verbose = FALSE
)
```

### Save and reload the model

A metric model is an ordinary R object. Save and reload it with the base R
functions `saveRDS()` and `readRDS()`:

```{r}
#| label: save-metric-model
metric_model_file <- tempfile(fileext = ".rds")
saveRDS(model, metric_model_file)
model <- readRDS(metric_model_file)
```

Keep the `.rds` file to apply the model in a future R session without retraining.

## Step 5: Classify new landscapes

Use `apply_metric_model()` to classify landscapes that were not used for metric
selection or training. These landscapes can serve two roles:

- *Labelled test landscapes* have known patterns and can be used to evaluate model
  performance
- *Application landscapes* have unknown patterns and can be
  classified with the fitted model

The function call is the same in both cases. This example uses a labelled,
independent test set so that both the predictions and model performance can be
inspected. Application landscapes could instead be imported data or output from
simulation models (see
[Import user-defined landscapes](importing-landscapes.html)).

:::{.callout-note}

## Landscape geometry

Training and application landscapes should have comparable extent, resolution,
and aspect ratio because a mismatched geometry can produce substantially different
metric values.
`apply_metric_model()` warns when it detects a substantial mismatch. See
[Matching training and application data](matching-training-application.html)
for details.

:::

```{r}
#| label: classify-landscapes
# Use a separate seed for independent test landscapes
set.seed(654321)

# Create 30 test landscapes, 10 from each training pattern
test_landscapes <- create_landscapes(
  n = 30,
  patterns = c("sharp", "clustered", "bands")
)
```

`apply_metric_model()` automatically calculates the metrics required by the model and applies
the centering and scaling stored during training. Because these artificial test
landscapes retain their true pattern labels, the default `evaluate = "auto"`
also evaluates the predictions. The same function call applies to unknown
landscapes, but no performance evaluation is returned.

:::{.callout-note}
## Optional performance evaluation

Performance is evaluated automatically when the input landscapes have known
pattern labels. Use `evaluate = "none"` to classify without evaluation, even
when the landscapes are labelled. Use `evaluate = "required"` to raise an error
when performance cannot be evaluated.
:::

```{r}
#| label: apply-model
#| message: false
#| warning: false
# Classify test landscapes using the trained model
classification <- apply_metric_model(
  landscapes = test_landscapes,
  model = model,
  verbose = FALSE
)
```

The result always contains a prediction table you can access with `$predictions`. This table lists the predicted pattern class, the predicted-class score, and one score for every class the
model learned. If available from the input data, it also gives the actual pattern label. `$performance` contains the evaluation results for labelled
landscapes and is `NULL` when the true patterns are unknown or evaluation is
disabled.

```{r}
#| label: view-predictions-2
# Predicted patterns
classification$predictions
```

The per-class scores are non-negative and sum to one, but they are not
calibrated probabilities. The `score` column contains the largest per-class
score for each landscape. You can compare the per-class scores within a row to
assess how decisive the model was: one dominant score indicates stronger
support for one class, while similar scores indicate ambiguity. A score of 0.8
does not mean that the prediction has an 80% probability of being correct.

For labelled test landscapes, use the performance results to inspect the
confusion matrix, overall model accuracy, and class-specific precision, recall, and
F1-score:

```{r}
#| label: view-test-performance
classification$performance$confusion_matrix
classification$performance$accuracy
classification$performance$per_class_metrics
```

Use `plot_classified_landscapes()` to show landscapes with their true and
predicted patterns. Correct classifications are blue and misclassifications are
bold orange. The example shows 12 of the 30 test landscapes to keep the figure
readable. Set `only_misclassified = TRUE` to inspect only misclassifications.

:::{.callout-note}
If you classified landscapes where the true patterns are not known, the plot will
show the landscape and the predicted pattern only.
:::

```{r}
#| label: plot-classified-landscapes
#| fig-width: 8
#| fig-height: 6
#| fig-alt: "Twelve test landscapes labelled with their actual and predicted ecotone pattern classes"
#| fig-cap: "Classification results for 12 independent test landscapes. Blue labels indicate correct predictions and bold orange labels indicate misclassifications."
plot_classified_landscapes(
  classification = classification$predictions,
  landscapes = test_landscapes,
  subset_index = 1:12,
  ncol = 4
)
```

## Optional: Estimate performance with cross-validation

The independent test set above evaluates the final classifier on new, labelled
landscapes. Cross-validation serves a different purpose: it assesses how
consistently a model classifies held-out subsets of the training data. This can
help when comparing model settings.

Cross-validation can also provide an estimate of model performance when too few
labelled landscapes are available for an independent test. Because the metrics
were selected using the full training set before cross-validation, the held-out
folds are not fully independent. The cross-validation results therefore assess
model fitting with the selected metrics, not the complete metric-selection and
training workflow.

The following code runs five-fold cross-validation and prints the resulting
performance summaries.

```{r}
#| label: optional-cross-validation
# Reset the seed immediately before cross-validation
set.seed(123456)

cv_model <- train_metric_model(
  metrics = landscape_metrics,
  metrics_selected = metric_selection,
  cv_method = "k-fold",
  cv_folds = 5,
  verbose = FALSE
)

cv_model$performance$confusion_matrix
cv_model$performance$accuracy
cv_model$performance$per_class_metrics
```

The performance summaries combine the held-out predictions from all folds.
Each landscape is classified once by a model that was not trained on that
landscape. The confusion matrix contains these pooled predictions, and the
accuracy and per-class metrics are calculated from that matrix. After cross-validation, the returned final model is
trained on all training landscapes.

For small or unbalanced training sets, `train_metric_model()` may reduce the
number of folds or switch to leave-one-out cross-validation. You can also
request leave-one-out cross-validation directly with `cv_method = "loo"`, which
trains one model per landscape and is therefore only feasible for very small
datasets. See `?train_metric_model` for more details.
