---
title: "Classify landscapes with pixels"
vignette: >
  %\VignetteIndexEntry{Classify landscapes with pixels}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
    cache: true
---

This guide shows how to train a classifier on categorical landscape rasters and
use it to classify new landscapes. The pixel-based workflow uses a
convolutional neural network (CNN) from the `keras3` package to learn directly
from raster-cell values.

The example uses the same three ecotone patterns as the quick example in
[Get started with patternscaper](patternscaper.html): sharp, clustered, and
bands.

```{r}
#| label: setup
library(patternscaper)
```

## Overview

The workflow has three steps:

1. Prepare labelled training landscapes with known patterns
2. Train the classifier on the landscape pixel data with `train_pixel_model()`
3. Classify new landscapes with `apply_pixel_model()` and evaluate the predictions when their true patterns are known

## Before you begin

### Set up Keras

Pixel models use `keras3` with a TensorFlow backend. Complete the
[installation guide](install-keras.html) before starting this workflow.

In a fresh R session, initialize Keras and verify which backend it is using:

```{r}
#| label: test-keras
keras3::config_backend()
#> [1] "tensorflow"
```

This must return `"tensorflow"` before the pixel workflow can run. The call
initializes Keras and Python, so it is an active verification that your setup works correctly. If it returns another backend or raises an error,
return to the Keras installation guide and look at `reticulate::py_config()` in the
failed session.

### Set seeds for reproducibility

Use `set.seed()` before generating landscape datasets and `set_random_seed()` immediately before each model fit. The first resets only R's random number generator (RNG), while the second resets both R's and Keras' RNG. Exact results can still vary slightly across hardware and software configurations.

## Step 1: Create training landscapes

Create training landscapes with known patterns. The
[artificial landscape guide](landscape-generation.html) explains how to create
training batches, and the [pattern gallery](pattern-gallery.html) shows the
available patterns and their parameters. Alternatively, import your own raster
or matrix data as described in
[Import user-defined landscapes](importing-landscapes.html). See
[Matching training and application data](matching-training-application.html)
for pixel-model input requirements.

```{r}
#| label: generate-landscapes
set.seed(1)
# Generate 100 training landscapes from 3 ecotone patterns
training_landscapes <- create_landscapes(
  n = 100,
  patterns = c("sharp", "clustered", "bands")
)
```

Training landscapes should represent the pattern classes, variation, and
ecological scale expected in the application data.

## Step 2: Train the model

`train_pixel_model()` treats the numeric land-cover codes as unordered
categories and converts each code into its own binary input channel.

The default `"multiscale"` CNN learns spatial features at different
neighbourhood sizes. You can adjust settings such as the number of epochs,
batch size, learning rate, and dropout. See `?train_pixel_model` for details.
If you are familiar with Keras, you can also supply a custom architecture (see
the [custom architecture section](#optional-use-a-custom-model-architecture)).

Here, we train one model on all training landscapes for 20 epochs.
Model training also supports
[early stopping](#optional-stop-training-early-with-validation-data) and
[cross-validation](#optional-estimate-performance-with-cross-validation).

```{r}
#| label: train-model
#| message: false
#| output: false
set_random_seed(2)
model <- train_pixel_model(
  landscapes = training_landscapes,
  cv_method = "none",
  epochs = 20,
  verbose = FALSE
)
```

### Save and reload the model

A pixel model combines a trained Keras network with R metadata. Save and reload
the complete classifier with `save_pixel_model()` and `load_pixel_model()`:

```{r}
#| label: save-pixel-model
pixel_model_bundle <- tempfile("pixel-model-")
save_pixel_model(model, pixel_model_bundle)
model <- load_pixel_model(pixel_model_bundle)
```

The model directory contains `model.keras` and `metadata.rds`. Keep the complete
directory to apply the model in a future R session without retraining. Set
`overwrite = TRUE` to replace an existing model bundle.

## Step 3: Classify new landscapes

Use `apply_pixel_model()` to classify landscapes that were not used for
training. These landscapes can serve two roles:

- *Labelled test landscapes* have known patterns and can be used to evaluate
  model performance
- *Application landscapes* have unknown patterns and can be classified with the
  fitted model

The function call is the same in both cases. This example uses a labelled,
independent test set so that both the predictions and model performance can be
inspected. Application landscapes could instead be imported data or simulation
output; see [Import user-defined landscapes](importing-landscapes.html).

:::{.callout-note}
## Pixel-model inputs

Application landscapes must contain one categorical raster layer with no `NA`
cells. They may omit land-cover codes used during training, but cannot introduce
new ones. `apply_pixel_model()` resizes rasters to the training dimensions and
warns when a different aspect ratio would stretch the pattern. Training and
application landscapes should also represent comparable ecological extents and
resolutions. See
[Matching training and application data](matching-training-application.html)
for details.
:::

```{r}
#| label: test-landscapes
# Use a separate seed for independent test landscapes
set.seed(4)

# Create 30 test landscapes, 10 from each training pattern
test_landscapes <- create_landscapes(
  n = 30,
  patterns = c("sharp", "clustered", "bands")
)
```

Because these artificial test landscapes retain their true pattern labels, the
default `evaluate = "auto"` also evaluates the predictions. The same function
call applies to landscapes with unknown patterns, but no performance evaluation
is returned.

:::{.callout-note}
## Optional performance evaluation

Performance is evaluated automatically when true patterns are known. Use
`evaluate = "none"` to classify without evaluation, even when the landscapes
carry pattern labels. Use `evaluate = "required"` to raise an error when
performance cannot be evaluated.
:::

```{r}
#| label: apply-model
#| output: false
# Classify test landscapes using the trained model
classification <- apply_pixel_model(
  landscapes = test_landscapes,
  model = model,
  verbose = FALSE
)
```

The result always contains a prediction table in `$predictions`. It lists the
predicted pattern class, the predicted-class score, and one score for every
class the model learned. If available from the input data, it also gives the
actual pattern label. `$performance` contains evaluation results for labelled
landscapes and is `NULL` otherwise.

```{r}
#| label: view-predictions-2
# Predicted patterns
classification$predictions
```

The per-class scores are non-negative and sum to one, but they are not
calibrated probabilities. The `score` column contains the largest per-class
score for each landscape. Compare the per-class scores within a row to assess
how decisive the model was: one dominant score indicates stronger support for
one class, while similar scores indicate ambiguity. A score of 0.8 does not mean
that the prediction has an 80% probability of being correct.

For labelled test landscapes, use the performance results to inspect the
confusion matrix, overall model accuracy, and class-specific precision, recall,
and F1-score:

```{r}
#| label: view-test-performance
classification$performance$confusion_matrix
classification$performance$accuracy
classification$performance$per_class_metrics
```

Use `plot_classified_landscapes()` to show landscapes with their true and
predicted patterns. Correct classifications are blue and misclassifications are
bold orange. The example shows 12 of the 30 test landscapes to keep the figure
readable. Set `only_misclassified = TRUE` to show only misclassifications.

:::{.callout-note}
If you classified landscapes where the true patterns are not known, the plot will
show the landscape and the predicted pattern only.
:::

```{r}
#| label: plot-classified-landscapes
#| fig-width: 8
#| fig-height: 6
#| fig-alt: "Twelve test landscapes labelled with their actual and predicted ecotone pattern classes"
#| fig-cap: "Classification results for 12 independent test landscapes. Blue labels indicate correct predictions and bold orange labels indicate misclassifications."
# Visualize true and predicted patterns
plot_classified_landscapes(
  classification = classification$predictions,
  landscapes = test_landscapes,
  subset_index = 1:12,
  ncol = 4
)
```

## Optional: Stop training early with validation data

You can use validation data to stop training when validation loss no longer
improves. The default callback stops after `patience` epochs without improvement
and then restores the model weights from the best epoch. This option requires
`cv_method = "none"`.

Early stopping can reduce overfitting to the training data, but the validation
set must be large enough and representative of the data the model will later
classify. It must contain every training pattern class and use the same raster
dimensions and land-cover codes. Here, we create a separate set of artificial
validation landscapes.

```{r}
#| label: optional-early-stopping
#| eval: false

# Create validation landscapes
set.seed(3)
validation_landscapes <- create_landscapes(
  n = 30,
  patterns = c("sharp", "clustered", "bands")
)

# Train with early stopping based on validation loss
set_random_seed(2)
early_stopped_model <- train_pixel_model(
  landscapes = training_landscapes,
  validation_landscapes = validation_landscapes,
  cv_method = "none",
  epochs = 100,
  patience = 15,
  verbose = FALSE
)
```

If a separate validation set is not available, `validation_split = 0.2`
creates a stratified split from the training landscapes. Validation results can
guide model development, but they do not replace evaluation on an untouched
test set.

## Optional: Estimate performance with cross-validation

The independent test set above evaluates the final classifier on new, labelled
landscapes. Cross-validation serves a different purpose: it assesses how
consistently a model classifies held-out subsets of the training data. This can
help when comparing model settings.

Cross-validation can also estimate model performance when too few labelled
landscapes are available for an independent test. If it is used to choose model
settings, its results are not a final independent evaluation.

The following code runs five-fold cross-validation. It is not executed when
this guide is rendered because pixel-model training is computationally
expensive.

```{r}
#| label: optional-cross-validation
#| eval: false

# Reset both random-number generators immediately before cross-validation
set_random_seed(2)
cv_model <- train_pixel_model(
  landscapes = training_landscapes,
  cv_method = "k-fold",
  cv_folds = 5,
  epochs = 20,
  verbose = FALSE
)

cv_model$performance$confusion_matrix
cv_model$performance$accuracy
cv_model$performance$per_class_metrics
```

The summaries combine the held-out predictions from all folds. Each landscape
is classified once by a model that was not trained on that landscape. After
cross-validation, the returned final model is trained on all training
landscapes.

Cross-validation folds run for the full number of epochs and cannot use
validation data. For small
or unbalanced training sets, `train_pixel_model()` may reduce the number of
folds or switch to leave-one-out cross-validation. Leave-one-out trains one
model per landscape and is only practical for very small datasets. See
`?train_pixel_model` for details.

## Optional: Use a custom model architecture

If you are familiar with Keras, you can pass a model-building function to the
`architecture` parameter.
The function must accept the four arguments shown below and return a new,
uncompiled Keras model. `train_pixel_model()` then compiles and trains it.

```{r}
#| label: optional-custom-architecture
#| eval: false

shallow_cnn <- function(input_shape, n_classes, dropout_rate, dense_units) {
  keras3::keras_model_sequential(input_shape = input_shape) |>
    keras3::layer_conv_2d(
      filters = 32,
      kernel_size = c(3, 3),
      activation = "relu"
    ) |>
    keras3::layer_max_pooling_2d() |>
    keras3::layer_flatten() |>
    keras3::layer_dropout(rate = dropout_rate) |>
    keras3::layer_dense(units = dense_units, activation = "relu") |>
    keras3::layer_dense(units = n_classes, activation = "softmax")
}

set_random_seed(2)
custom_model <- train_pixel_model(
  landscapes = training_landscapes,
  architecture = shallow_cnn,
  cv_method = "none",
  epochs = 20,
  verbose = FALSE
)
```

The final layer must contain one softmax output for each pattern class. The
`n_classes` argument supplies this number automatically. Use the built-in
architecture unless you are comfortable designing and evaluating neural
networks.
