---
title: "Classify landscapes using landscape metrics"
vignette: >
  %\VignetteIndexEntry{classify-metrics}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
    cache: true
---

This vignette shows how to train a neural network on landscape metrics and use
it to classify new landscapes.

```{r}
#| label: setup
library(patternscaper)

# Set seed for reproducibility
set.seed(123456)
```


The workflow has five steps:

1. Generate labelled training landscapes with known patterns using `create_landscapes()`
2. Calculate their landscape metrics with `calculate_metrics()`
3. Select informative metrics with `evaluate_metrics()`
4. Train a neural network using the selected metrcis with `train_metric_model()`
5. Classify new landscapes with `apply_metric_model()`


## Step 1: Generate Training Landscapes

Generate training landscapes with known patterns. The
[landscape generation vignette](landscape-generation.html) describes all
available patterns and options.

```{r}
#| label: generate-landscapes
# Generate 100 training landscapes of 3 patterns
landscapes <- create_landscapes(
  n = 100,
  patterns = c("labyrinth", "random", "clustered")
)
```

## Step 2: Calculate Landscape Metrics

Next, you can calculate landscape metrics for the training landscapes. 
Landscape metrics are calculated using the metrics and functions from the [`landscapemetrics`](https://r-spatialecology.github.io/landscapemetrics/) package.
They can be calculated on the class level (i.e. for each land cover class separately) or on the landscape level (i.e. for the whole landscape).
For details, see [landscape metrics vignette](landscape-metrics.html).

In this example, we calculate metrics on the landscape level.


```{r}
#| label: calculate-metrics
#| message: false
# Calculate landscape metrics on the landscape level
landscape_metrics <- calculate_metrics(
  landscapes,
  level = "landscape"
)
```

## Step 3: Evaluate Landscape Metrics

Use `evaluate_metrics()` to rank and select a requested number of metrics. There are different ranking methods available (e.g. Kruskal-Wallis H test). The
[landscape metrics vignette](landscape-metrics.html) describes the other
methods.

:::{.callout-note}
Some metrics are `NA` or unavailable for particular landscapes. At the class
level, for example, a metric cannot be calculated for a class absent from a
landscape. Other metrics may have the same value for all landscapes and therefore no variation. By default,
`evaluate_metrics()` removes these metrics and reports which were excluded.

It is recommended to leave `exclude_incomplete_metrics = TRUE` (the default), as metrics that are missing for some landscapes cannot be used for training the neural network. They may however be informative and useful for other types of analyses.
:::


```{r}
#| label: evaluate-metrics
best_10 <- evaluate_metrics(
  metrics = landscape_metrics,
  method = "kruskal_effsize",
  metrics_number = 10,
  verbose = FALSE
)

best_10
```

`evaluate_metrics()` returns the selected names in `$selected` together with the
full ranking in `$ranking`, which records the score of every metric and what
happened to the ones that were not selected. See the
[landscape metrics vignette](landscape-metrics.html) for how to read it. The
functions below accept either the object itself or `best_10$selected`.

Alternatively, provide a character vector of metric names to select them
manually.

You can plot the differences between patterns for the selected metrics using the `plot_metrics` function.

```{r}
#| label: plot-metrics
#| fig-width: 8
#| fig-height: 6
#| fig-cap: "Boxplots show the distribution of the selected landscape metrics across different spatial patterns. Each point represents one training landscape and each panel is one of the selected metrics."
# Plot selected metrics
plot_metrics(
  metrics = landscape_metrics,
  selected_metrics = best_10
)
```

## Step 4: Train Neural Network

The model can now be trained using the selected metrics from step 3.
Training can be done with or without cross-validation.
Valid options for the cross-validation method `cv_method` are:

- `"none"`: no cross-validation
- `"k-fold"`: k-fold cross-validation with `cv_folds` number of folds
- `"loo"`: leave-one-out cross-validation

See `?train_metric_model` for further options.

```{r}
#| label: train-model
# Train neural network with k-fold cross-validation (3 folds)
model <- train_metric_model(
  metrics = landscape_metrics,
  metrics_selected = best_10,
  cv_method = "k-fold",
  cv_folds = 3,
  verbose = FALSE
)
```

To check the model performance, you can look at the confusion matrix from cross-validation:

```{r}
#| label: view-confusion-matrix
# Confusion matrix from cross-validation
model$performance$confusion_matrix
```

You can also check other performance metrics like overall accuracy:

```{r}
#| label: view-overall-accuracy
# Overall accuracy
model$performance$accuracy
```

Or per class metrics like precision, recall, and F1-score:

```{r}
#| label: view-per-class-metrics
model$performance$per_class_metrics
```

## Save and Reload the Model

A metric model is an ordinary R object. Save and reload it with the base R
functions `saveRDS()` and `readRDS()`:

```{r}
#| label: save-metric-model
metric_model_file <- tempfile(fileext = ".rds")
saveRDS(model, metric_model_file)
model <- readRDS(metric_model_file)
```

Keep the `.rds` file to apply the model in a future R session without retraining.

## Step 5: Classify New Landscapes

Create test landscapes and classify them with the trained model. These could
instead be imported data or output from simulation models; see the
[importing landscapes vignette](importing-landscapes.html).

```{r}
#| label: classify-landscapes
# Create 10 new test landscapes using the same patterns as for training
test_landscapes <- create_landscapes(
  n = 10,
  patterns = c("labyrinth", "random", "clustered")
)
```

If test landscapes have been created with `create_landscapes()`, their true patterns are known. Therefore they can be used to evaluate classification performance.

The result always has the same two parts, `$predictions` and `$performance`.

:::{.callout-note}
Performance is evaluated automatically when true patterns are known. Use
`evaluate = "none"` to classify without evaluation, even when the landscapes
carry pattern labels. Use `evaluate = "required"` to raise an error when
performance cannot be evaluated.
:::

```{r}
#| label: apply-model
#| message: false
#| warning: false
# Classify test landscapes using the trained model
classification <- apply_metric_model(
  landscapes = test_landscapes,
  model = model
)
```

You can look at the predicted patterns for each test landscape individually and
compare actual and predicted classes:

```{r}
#| label: view-predictions-2
# Predicted patterns
classification$predictions
```

You can also look at performance summaries like confusion matrix:

```{r}
#| label: view-performance-2
# Performance summary
classification$performance$confusion_matrix
```

And other metrics like accuracy, precision, recall, and F1-score:

```{r}
#| label: view-per-class-metrics-2
# Other performance metrics
classification$performance$per_class_metrics
```

Use `plot_classified_landscapes()` to show landscapes with their true and
predicted patterns. Correct classifications are blue and misclassifications are
bold orange. Set `only_misclassified = TRUE` to show only misclassifications.

:::{.callout-note}
If you classified landscapes where the true patterns are not known, the plot will
show the landscape and the predicted pattern only.
:::

```{r}
#| label: plot-classified-landscapes
#| fig-width: 8
#| fig-height: 6
# Visualize true and predicted patterns

plot_classified_landscapes(
  classification = classification$predictions,
  landscapes = test_landscapes
)
```
