---
title: "Classify landscapes using Keras on landscape rasters"
vignette: >
  %\VignetteIndexEntry{classify-landscapes-keras}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
    cache: true
---

This vignette shows how to classify landscape patterns using a convolutional neural network (Keras backend) trained directly on landscape raster data.

```{r}
#| label: setup
library(spatPatClassifyR)
```

The workflow consists of the following steps:

1. Generate training landscapes with known patterns using the `create_landscapes()` function.
2. Train a neural network on the landscape pixel data using the `train_pixel_model()` function.
3. Classify new landscapes using the trained model with the `apply_pixel_model()` function.

## Setup Keras

Models are trained using the R package `keras3`. It requires a working installation of TensorFlow.

You can refer to our [installation guide](install-keras.html) and the official [`keras3` package website](https://keras3.posit.co/) for instructions on how to install Keras and TensorFlow.

Use the package's backend check to confirm that `keras3` can reach TensorFlow:

```{r}
#| label: test-keras
keras_available()
```

This must return `TRUE` before the pixel workflow can run.

## Set seed for reproducibility

The pixel workflow uses R's random-number generator for operations such as
cross-validation fold assignment and data shuffling. Keras and TensorFlow use
their own randomness for weight initialization and dropout. Use `set.seed()` to
reproduce landscape generation, and call `set_random_seed()` immediately before
training to reset both R and Keras.

```{r}
#| label: set-seed
set.seed(123456)
```

## Step 1: Generate Training Landscapes

You can generate a set of training landscapes with known patterns.
See [landscape generation vignette](landscape-generation.html) for details on landscape generation and available patterns and options.

```{r}
#| label: generate-landscapes
training_landscapes <- create_landscapes(
  n = 100,
  patterns = c("labyrinth", "random", "clustered")
)
```

## Step 2: Train Neural Network

The model can be trained with or without cross-validation. Valid methods for 
the cross-validation method `cv_method` are:

- `"none"`: No cross-validation, train on all data.
- `"k-fold"`: k-fold cross-validation with `cv_folds` number of folds.
- `"loo"`: leave-one-out cross-validation.

For details and further options see the function help of `train_pixel_model()`.

The cell values must be numeric whole-number codes for unordered habitat
categories. For example, `0` might represent bare ground and `1` vegetation, or
`1`, `2`, and `3` might represent forest, grassland, and water. The numerical
size and order of these codes have no meaning. `train_pixel_model()` converts
each habitat value into its own binary input channel. A raster with three
habitat values therefore produces three channels. Text labels are not
supported.

The built-in architecture, `"multiscale"`, processes the landscapes as raster
arrays. It has two sequential convolutional blocks. Each block applies 3 by 3 and
5 by 5 convolutions followed by max pooling, with 32 filters in the first block and
64 in the second. Dropout and a dense layer precede the softmax output. Users can
tune the learning rate, epochs, batch size, dropout, dense-layer size, and training
validation settings; see `?train_pixel_model` for the complete interface.
Experienced Keras users can also supply a custom architecture; see the optional
section at the end of this guide.

Here, we train the model using 2-fold cross-validation to keep computational time
low. Every fold trains for the requested number of epochs, and its held-out
landscapes are then used to estimate performance. After cross-validation, the
function trains a fresh final model on all supplied landscapes.

:::{.callout-note}
## Low fold accuracy

The function will print fold accuracies during training (see below). If the accuracy is low,
consider increasing the number of training landscapes or check the function help for other options to improve model performance.
:::

```{r}
#| label: train-model
#| message: false
#| output: false
set_random_seed(123456)
model <- train_pixel_model(
  landscapes = training_landscapes,
  cv_method = "k-fold",
  cv_folds = 2,
  verbose = FALSE
)
```

To check the model performance, we can look at the confusion matrix from cross-validation:

```{r}
#| label: view-confusion-matrix
# Confusion matrix from cross-validation
model$performance$confusion_matrix
```

You can also check other performance metrics like overall accuracy:

```{r}
#| label: view-overall-accuracy
# Overall accuracy
model$performance$accuracy
```

Or per class metrics like precision, recall, and F1-score:

```{r}
#| label: view-per-class-metrics
model$performance$per_class_metrics
```

### Optional: Use Validation Data for Early Stopping

For a final model without cross-validation, validation landscapes can be used
to monitor how well the model performs on data that are not used to train the model. With the default callback, training stops when validation loss has not
improved for `patience` epochs, and the weights from the best epoch are restored.

When enough data are available, create or set aside a separate, labelled
validation set. This validation set must contain every training pattern class and use the same raster dimensions and habitat codes:

```r
validation_landscapes <- create_landscapes(
  n = 30,
  patterns = c("labyrinth", "random", "clustered")
)

set_random_seed(123456)
model <- train_pixel_model(
  landscapes = training_landscapes,
  validation_landscapes = validation_landscapes,
  cv_method = "none",
  epochs = 100,
  patience = 15
)
```

If the available landscapes cannot support a separate validation set, request
a stratified split instead. For example, `validation_split = 0.2` assigns about
20% of every pattern class to validation and fits the model weights with the
remaining landscapes.

The returned `performance` contains information on the validation loss, accuracy, per-class
metrics, the selected epoch, and the validation predictions. These results help
choose the training procedure, but they are not final evidence of model
performance because the validation data influenced model development. Use
`apply_pixel_model()` with a separate test set for the final evaluation.

## Save and Reload the Model

A pixel model contains both a Keras network and R metadata. Save the complete
classifier with
`save_pixel_model()` and reload it with `load_pixel_model()`:

```{r}
#| label: save-pixel-model
pixel_model_bundle <- tempfile("pixel-model-")
save_pixel_model(model, pixel_model_bundle)
model <- load_pixel_model(pixel_model_bundle)
```

The supplied path is a directory containing `model.keras` and `metadata.rds`.
Treat the complete directory as one model: move, copy, or archive the directory
rather than the single files inside it. If a model already exists it cannot be overwritten, unless `overwrite = TRUE` is set.

## Step 3: Classify New Landscapes

Finally, you can create some new test landscapes and classify them using the trained model. In this example, we create 20 new landscapes for testing. 

In reality this could be landscapes read in from files or created in other ways.
For details on importing own landscapes, see the [importing landscapes vignette](importing-landscapes.html).

Application landscapes must contain one raster layer with numeric categorical
values and have no `NA` cells. They must use the same category codes as the
training data. A trained habitat category may be absent from a new landscape,
but a new code that was not present during training causes an error.
`apply_pixel_model()` resamples rows and columns to the training dimensions with
nearest-neighbor resampling. It warns when a different aspect ratio means that
this stretches the pattern. See
[Matching training and application data](matching-training-application.html) for
the practical consequences of resizing.

```{r}
#| label: test-landscapes
test_landscapes <- create_landscapes(
  n = 20,
  patterns = c("labyrinth", "random", "clustered")
)
```

If test landscapes have been created with `create_landscapes()`, their true patterns are known. Therefore they can be used to evaluate classification performance.

The result always has the same two parts, `$predictions` and `$performance`.

:::{.callout-note}
Performance is evaluated automatically when the true patterns are known. Landscapes that carry a `pattern` but you don't want to evaluate model performance with them but only classify them can skip evaluation with `evaluate = "none"`. Setting `evaluate = "required"` raises an error instead of quietly returning no performance when the true patterns turn out to be missing in some or all cases.
:::


```{r}
#| label: apply-model
#| output: false
classification <- apply_pixel_model(
  landscapes = test_landscapes,
  nn_model = model
)
```

You can look at the predicted patterns for each test landscape individually and
compare actual and predicted classes:

```{r}
#| label: view-predictions-2
# Predicted patterns
classification$predictions
```

The `predictions` table contains the predicted class, its `score`, and one score
for every trained class. The class scores are non-negative and sum to one, but
they are not calibrated probabilities. A score of 0.8 means that the model
assigned 80% of its total output for this landscape to the predicted class and
20% to all other classes combined. It does not mean that the prediction has an
80% probability of being correct. Use the scores to judge how decisive the model
was. A dominant score indicates stronger support for one class, while similar
scores for several classes indicate that the landscape was ambiguous to the
model.

You can also look at performance summaries like confusion matrix:

```{r}
#| label: view-performance-2
# Performance summary
classification$performance$confusion_matrix
```

And other metrics like accuracy, precision, recall, and F1-score:

```{r}
#| label: view-per-class-metrics-2
# Other performance metrics
classification$performance$per_class_metrics
```

To visualize the classified landscapes along with their true and predicted patterns
use the function `plot_classified_landscapes`. Correctly classified landscapes are shown in blue, misclassified ones in bold orange. To plot only misclassified landscapes, set `only_misclassified = TRUE`.

:::{.callout-note}
If you classified landscapes where the true patterns are not known, the plot will
show the landscape and the predicted pattern only.
:::

```{r}
#| label: plot-classified-landscapes
#| fig-width: 8
#| fig-height: 8
# Visualize true and predicted patterns

plot_classified_landscapes(
  classification = classification$predictions,
  landscapes = test_landscapes
)
```

## Optional: Use a Custom Architecture

Users familiar with Keras can pass a model-building function to `architecture`.
The function must accept the four arguments shown below and return a new,
uncompiled Keras model. `train_pixel_model()` calls it separately for each
cross-validation fold and for the final model, then compiles and trains each
model.

```r
shallow_cnn <- function(input_shape, n_classes, dropout_rate, dense_units) {
  keras3::keras_model_sequential(input_shape = input_shape) |>
    keras3::layer_conv_2d(
      filters = 32,
      kernel_size = c(3, 3),
      activation = "relu"
    ) |>
    keras3::layer_max_pooling_2d() |>
    keras3::layer_flatten() |>
    keras3::layer_dropout(rate = dropout_rate) |>
    keras3::layer_dense(units = dense_units, activation = "relu") |>
    keras3::layer_dense(units = n_classes, activation = "softmax")
}

custom_model <- train_pixel_model(
  landscapes = training_landscapes,
  architecture = shallow_cnn
)
```

The final layer must contain one softmax output for each pattern class. The
`n_classes` argument supplies this number automatically. The built-in architecture
is recommended unless the user is comfortable designing and evaluating neural
networks.
