---
title: "importing-landscapes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{importing-landscapes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(spatPatClassifyR)
```

All functions in `spatPatClassifyR` expect landscapes to be a `landscape` object or a list of `landscape` objects. 
If you want to train or classify with landscapes that you have created or read in from a file, you will need to convert them to `landscape` objects first. This vignette shows how to do that.

Please note that right now, we only support categorical landscapes.

To convert your own landscapes to `landscape` objects, you can use the `landscape()` function

The function excepts landscape data either as a matrix or a a `SpatRaster`.

You can manually set `pattern`, `name` and `params` metadata fields.

### From a matrix

```{r }
# Create a sample matrix
landscape_matrix <- matrix(
  data = sample(1:3, 100, replace = TRUE),
  nrow = 10,
  ncol = 10
)

# Convert the matrix to a landscape object
landscape_object <- landscape(
  landscape_matrix,
  name = "Sample Landscape",
  pattern = "custom"
)
landscape_object
```

### From a SpatRaster

Here, we use the `terra` package to create a `SpatRaster` from the matrix and then convert it to a `landscape` object. But this could also be a raster read in from a file.

```{r }
library(terra)
# Create a sample SpatRaster
landscape_raster <- rast(landscape_matrix)
# Convert the SpatRaster to a landscape object
landscape_object_raster <- landscape(
  landscape_raster,
  name = "Sample Raster Landscape",
  pattern = "custom"
)
landscape_object_raster
```

```{r}
plot_landscape_list(
  list(landscape_object, landscape_object_raster),
  titles = "name"
)
```

## Converting multiple landscapes

If you have multiple landscapes to convert, you can use use the map function from the `purrr` package to apply the `landscape()` function to each landscape in a list. 

If you have for example a list of matrices, you batch convert them and optionally set the `name` and `pattern` for each landscape like this:


```{r}
library(purrr)
# Create a list of sample matrices
landscape_matrices <- list(
  matrix1 = matrix(sample(1:3, 100, replace = TRUE), nrow = 10),
  matrix2 = matrix(sample(1:3, 100, replace = TRUE), nrow = 10),
  matrix3 = matrix(sample(1:3, 100, replace = TRUE), nrow = 10)
)
# Define a vector with landscape names for each landscape
landscape_names <- c("Landscape 1", "Landscape 2", "Landscape 3")
# Define a vector with landscape patterns for each landscape
landscape_patterns <- c("pattern_a", "pattern_b", "pattern_c")

# Convert the list of matrices to a list of landscape objects
landscape_objects <- pmap(
  list(
    landscape_matrices,
    landscape_names,
    landscape_patterns
  ),
  function(mat, name, pattern) {
    landscape(mat, name = name, pattern = pattern)
  }
)
# Plot the landscapes with their names and patterns as titles
plot_landscape_list(landscape_objects, titles = "both")
```
