---
title: "Import user-defined landscapes"
vignette: >
  %\VignetteIndexEntry{importing-landscapes}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

Convert your own categorical land-cover matrices or rasters into `landscape`
objects for model training or classification. If you are new to the package,
begin with [Get started with patternscaper](patternscaper.html) for an overview
of the complete classification workflow.

Input data can be created manually or read from files.

```{r}
#| label: setup
library(patternscaper)

set.seed(123456)
```

`patternscaper` functions that use raster landscapes all expect a `landscape` object or a list of `landscape` objects. To wrap a matrix or a `SpatRaster` from the [`terra` R package](https://rspatial.github.io/terra/index.html) into a `landscape` object, use the `landscape()` function.

Both workflows use categorical land-cover data represented by finite
whole-number codes. These codes don't need to start at zero or be consecutive.
This means that you should classify continuous measurements (e.g. elevation, vegetation indices), and raw images into land-cover classes before converting them to `landscape` objects. Use one categorical land-cover layer per landscape.

The pixel workflow does not support `NA` cells in the raster. The metric workflow
excludes `NA` cells from the analysed landscape. See
[Matching training and application data](matching-training-application.html)
for how missing cells and raster geometry affect the two workflows.

## Read or create input data

Read a classified raster file as a `SpatRaster` with `terra::rast()`:

```{r}
#| label: read-raster-file
#| eval: false
land_cover_raster <- terra::rast("path/to/classified-land-cover.tif")
```

Alternatively, you can create landscapes directly as a matrix. The examples below create a matrix and a `SpatRaster` containing the same
land-cover values:

```{r}
#| label: create-sample-landscapes
landscape_matrix <- matrix(
  sample(1:3, 100, replace = TRUE),
  nrow = 10
)

landscape_raster <- terra::rast(landscape_matrix)
```

For example, the values `1`, `2`, and `3` in the matrix can represent land-cover classes forest, grassland, and water. Training and application rasters must use the same code for the same land-cover category.

## Convert one landscape into a `landscape` object

Use `landscape()` to wrap the input data and optionally add metadata:

- `name` identifies an individual landscape in tables and plots.
- `pattern` is the known spatial pattern class used for model training and
  performance evaluation.

Both default to `NA` and can be set during conversion.

:::{.callout-note}
For model training, landscapes need a known pattern label. For application landscapes that should be classified, you can leave `pattern` at its
default `NA`.
:::

```{r}
#| label: convert-one-landscape
matrix_landscape <- landscape(
  landscape_matrix,
  name = "Matrix landscape example",
  pattern = "custom"
)

raster_landscape <- landscape(
  landscape_raster,
  name = "Raster landscape example",
  pattern = "custom"
)
```

Wrapping a `SpatRaster` in a `landscape` object preserves its cell values, extent, resolution, and coordinate reference system. Data without this spatial information (e.g. from a simple matrix) give a raster with a default cell resolution of 1 and
no coordinate reference system.

Printing a landscape object shows its properties:

```{r}
#| label: print-landscape
matrix_landscape
```

## Update landscape metadata

Names and known pattern labels can also be updated after conversion with
`set_landscape_name()` and `set_landscape_pattern()`:

```{r}
#| label: update-metadata
matrix_landscape <- matrix_landscape |>
  set_landscape_name("Landscape 1") |>
  set_landscape_pattern("pattern_a")
```

These functions change only the stored metadata. They do not alter the raster
cells.

## Convert multiple landscapes

Use `purrr::map()` to convert a list of matrices or rasters. Use `purrr::pmap()`
when names and patterns are supplied as parallel vectors.

```{r}
#| label: bulk-convert
landscape_matrices <- list(
  matrix1 = matrix(sample(1:3, 100, replace = TRUE), nrow = 10),
  matrix2 = matrix(sample(1:3, 100, replace = TRUE), nrow = 10),
  matrix3 = matrix(sample(1:3, 100, replace = TRUE), nrow = 10)
)
landscape_names <- c("Landscape 1", "Landscape 2", "Landscape 3")
landscape_patterns <- c("pattern_a", "pattern_a", "pattern_b")

landscape_objects <- purrr::pmap(
  list(
    data = landscape_matrices,
    name = landscape_names,
    pattern = landscape_patterns
  ),
  landscape
)
```

After conversion, you can use the landscapes like any other `landscape` objects in
`patternscaper`. For example, you can plot them with `plot_landscapes()`:

```{r}
#| label: plot-landscapes
#| fig-width: 8
#| fig-height: 3
#| fig-alt: "Three categorical landscapes labelled pattern_a, pattern_a, and pattern_b"
plot_landscapes(landscape_objects)
```

## Next steps

Continue with the guide for one of the classification workflows:

- [Classify landscapes with landscape metrics](classify-metrics.html)
- [Classify landscapes with pixels](classify-pixels.html)
