---
title: "Import user-defined landscapes"
vignette: >
  %\VignetteIndexEntry{importing-landscapes}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

Import matrices or rasters as training or classification data. These can be
created manually or read from files, such as classified remote-sensing images or model output.

The pixel workflow accepts one raster layer with categorical cover classes
represented by numeric whole-number codes. Text labels are not supported. The
codes need not start at zero or be consecutive.

```{r}
#| label: setup
library(patternscaper)
```

All functions in `patternscaper` expect landscapes to be a `landscape` object or a list of `landscape` objects. You can convert your own landscapes to `landscape` objects using the `landscape()` function.

You can convert landscapes from a matrix or a `SpatRaster` (from the
[`terra` R package](https://rspatial.github.io/terra/index.html)).

```{r}
#| label: create-sample-landscapes
# Create a sample matrix
landscape_matrix <- matrix(
  data = sample(1:3, 100, replace = TRUE),
  nrow = 10,
  ncol = 10
)

# Create a sample SpatRaster
landscape_raster <- terra::rast(landscape_matrix)
```

For example, `1`, `2`, and `3` can represent forest, grassland, and water. The
pixel workflow treats these codes as categories rather than quantities by
converting each value to a separate binary channel. Training and application
rasters must use the same code for the same land-cover category.

The resulting object stores the raster with its name and pattern metadata. Both
default to `NA` and can be set during conversion. When printed, missing values
are shown as `"unnamed"` and `"unclassified"`.

:::{.callout-note}
A known pattern label is needed for model training, but not for classification.
Leave an unknown pattern as `NA`. `"unclassified"` is also treated as unknown.
:::

```{r }
#| label: convert-matrix
# Convert the matrix to a landscape object
landscape_object_matrix <- landscape(
  landscape_matrix,
  name = "Sample Matrix Landscape",
  pattern = "custom"
)
# Convert the SpatRaster to a landscape object
landscape_object_raster <- landscape(
  landscape_raster,
  name = "Sample Raster Landscape",
  pattern = "custom"
)
```

Printing a landscape object shows its properties:

```{r}
#| label: print-landscape
landscape_object_matrix
```

## Update landscape metadata

Names and known pattern labels can also be updated after conversion with
`set_landscape_name()` and `set_landscape_pattern()`:

```{r}
#| label: update-metadata
landscape_object_matrix <- landscape_object_matrix |>
  set_landscape_name("Landscape 1") |>
  set_landscape_pattern("pattern_a")
```

These functions change only the stored metadata. They do not alter the raster
cells.

## Convert multiple landscapes

Use `lapply()` or `purrr::pmap()` to convert a
list of landscapes. `pmap()` is useful when names and patterns are supplied as
parallel vectors.

```{r}
#| label: bulk-convert
library(purrr)
# Create a list of sample matrices
landscape_matrices <- list(
  matrix1 = matrix(sample(1:3, 100, replace = TRUE), nrow = 10),
  matrix2 = matrix(sample(1:3, 100, replace = TRUE), nrow = 10),
  matrix3 = matrix(sample(1:3, 100, replace = TRUE), nrow = 10)
)
# Define a vector with landscape names for each landscape
landscape_names <- c("Landscape 1", "Landscape 2", "Landscape 3")
# Define a vector with landscape patterns for each landscape
landscape_patterns <- c("pattern_a", "pattern_b", "pattern_c")

# Convert the list of matrices to a list of landscape objects
landscape_objects <- pmap(
  list(
    landscape_matrices,
    landscape_names,
    landscape_patterns
  ),
  function(mat, name, pattern) {
    landscape(mat, name = name, pattern = pattern)
  }
)
```

After converting the landscapes, you can use them like any other landscape object in `patternscaper`.
You can for example use the plotting functions to visualize them:

```{r}
#| label: plot-landscapes
#| fig-width: 8
#| fig-height: 3
plot_landscapes(landscape_objects)
```
