---
title: "Install Keras and TensorFlow for R"
vignette: >
  %\VignetteIndexEntry{install-keras}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute: 
  eval: false
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

The workflow to classify landscapes directly from raster data using neural networks (see [vignette](classify-pixels.html)) requires `keras3` and its correct configuration.

In our experience, it can be a bit tricky to set up correctly, therefore we 
provide this guide to help you with the installation.

The setup requires the installation of the [`keras3` package](https://keras3.posit.co/) and the [`reticulate` package](https://rstudio.github.io/reticulate/), which provides an interface to Python from R.

The `keras3` package is installed together with `patternscaper`, and `reticulate` comes with it as a dependency of `keras3`, so you only need to load them:

```{r}
#| label: load-packages
library(patternscaper)
library(keras3)
library(reticulate)
```

First, you can check whether `keras3` can already reach a working TensorFlow
backend:

```{r}
#| label: check-keras
keras_available()
```

If this returns `TRUE`, your installation is already working and you can skip the rest of this guide.

## Specify Python Installation

Before installing Keras, you should specify which Python installation to use.

**Step 1: Find your Python installations**

Open a terminal (Command Prompt on Windows, Terminal on macOS/Linux) and run:

```bash
where python    # Windows
which python    # macOS/Linux
```

This will list all Python installations on your system.

**Step 2: Tell reticulate which Python to use**

Choose Python **3.10 or 3.11** (recommended for TensorFlow compatibility) and specify it:

```{r}
#| label: specify-python
# Replace this path with your actual Python 3.10 or 3.11 installation
reticulate::use_python(
  "C:/Users/YourName/AppData/Local/Programs/Python/Python310/python.exe",
  required = TRUE
)

# Verify reticulate found your Python
reticulate::py_discover_config()
```

This should display your Python installation details, not `NULL`.

:::{.callout-warning}
**Avoid Python 3.13+** as TensorFlow does not fully support these newer versions yet. Use Python 3.10 or 3.11.
:::

## Install Keras and TensorFlow

Now install Keras using the `install_keras()` function from the `keras3` package:

```{r}
#| label: install-keras
#| eval: false
install_keras()
```

This will take a while (5-15 minutes) and will install all necessary components in a virtual environment.

Now, check again if the installation was successful:

```{r}
#| label: check-keras-again
keras_available()
```

## Test Your Installation

To test if Keras is working correctly, you can run the following simple function:

:::{.callout-note}
The first time you run Keras functions after installation, TensorFlow needs to initialize, which may show some informational messages. Subsequent runs will be fast.
:::

```{r}
#| label: test-keras
keras3::to_categorical(0)
```

If this prints a matrix without errors, your Keras installation is working correctly.
