---
title: "Set up Keras and TensorFlow"
vignette: >
  %\VignetteIndexEntry{Set up Keras and TensorFlow}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  eval: false
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

The pixel-based workflow uses `keras3` with a TensorFlow backend. Installing `patternscaper` also installs the R package `keras3`, but the Python environment and TensorFlow backend need a
separate setup step.

This guide covers only what is needed to run the
[pixel-based workflow](classify-pixels.html). For platform-specific options, including GPU support, see the official [`keras3` installation reference](https://keras3.posit.co/reference/install_keras.html).

## Standard setup

Start a fresh R session, then run:

```{r}
#| label: install-keras
keras3::install_keras(backend = "tensorflow")
```

This creates a Python virtual environment named `r-keras` and
installs Keras, TensorFlow, and their Python dependencies. The download and installation can take several minutes.

When the installation finished, restart R.

## Verify the backend

In the fresh R session, load `patternscaper` and then ask Keras which backend it is using:

```{r}
#| label: verify-keras
library(patternscaper)
keras3::config_backend()
#> [1] "tensorflow"
```

The result of this call should be `"tensorflow"`. This step is an
active verification because `config_backend()` initializes Keras and Python, and `reticulate` resolves declared Python requirements as part of that process.

Tensorflow may print other messages into the console in that step. But as long as `config_backend()` returns `"tensorflow"`, the step was successful.

## Alternative: Set Python explicitly

The standard setup lets `keras3` choose a compatible Python installation automatically. If
the setup fails, or if you want to use a particular Python installation, you can select it manually before running the Keras installer.
Do this in a new R session before loading `patternscaper` or `keras3`.

:::{.callout-warning}
**Check compatibility before selecting Python manually.** Supported Python
versions depend on the current `keras3` installer and TensorFlow release. If
you do not require a particular Python installation, use the standard setup
and let `keras3` select one.
:::

First, locate the path of your desired Python executable. You can also find your Python installation from the terminal:

```bash
where python    # Windows
which python    # macOS or Linux
```

In your R session, set the Python path:

```{r}
#| label: specify-python
# Replace the path with your actual Python installation
reticulate::use_python(
  "C:/path/to/python.exe",
  required = TRUE
)
# Verify it worked
# Should print your python installation not NULL
reticulate::py_discover_config()
# Install tensorflow
keras3::install_keras(backend = "tensorflow")
```

Restart R after the installation finished, then repeat the
[backend verification](#verify-the-backend). Python compatibility changes over
time, so check the current
[TensorFlow installation guide](https://www.tensorflow.org/install/pip) before
selecting a version yourself. See the official
[`reticulate` Python selection reference](https://rstudio.github.io/reticulate/reference/use_python.html)
for other selection methods.

## Getting more help

If backend verification still fails after trying the standard setup or, if
needed, the explicit Python setup, consult the official
[Keras installation guide](https://keras.io/getting_started/) and the
[`keras3` installation reference](https://keras3.posit.co/reference/install_keras.html) for current setup instructions.

It can also be helpful to inspect your current python configuration:

```{r}
#| label: inspect-python
reticulate::py_config()
```

Once verification succeeds, continue with
[Classify landscapes using raster cells](classify-pixels.html).
