---
title: "Install Keras and TensorFlow for R"
vignette: >
  %\VignetteIndexEntry{install-keras}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

The workflow to classify landscapes directly from raster data using neural networks (see [vignette](classify-pixels.qmd)) `keras3` and its correct configuration.

In our experience, it can be a bit tricky to set up correctly, therefore we 
provide this guide to help you with the installation.

The setup requires the installation of the [`keras3` package](https://keras3.posit.co/) and the [`reticulate` package](https://rstudio.github.io/reticulate/), which provides an interface to Python from R.

Both packages are installed alongside the `spatPatClassifyR` package, so you only need to load them:

```{r}
#| label: load-packages
library(keras3)
library(reticulate)
```

First, you can check if you already have the keras and tensorflow modules installed
using:

```{r}
#| label: check-keras
reticulate::py_module_available("keras") &&
  reticulate::py_module_available("tensorflow")
```

If this returns `TRUE`, your installation is already working and you can skip the rest of this guide.

If it returns `FALSE`, you need to install Keras using the `install_keras()` function from the `keras3` package:

```{r}
#| label: install-keras
install_keras()
```

This will take a while (5-15 minutes) and should install all necessary components.

Now, check again if the installation was successful using the code from above.


```{r}
#| label: check-keras-again
reticulate::py_module_available("keras") &&
  reticulate::py_module_available("tensorflow")
```

To test if Keras is working correctly, you can run the following simple function from the `keras3` package.

:::{.callout-note}
If you run this for the first time, it might take a while as TensorFlow needs to be initialized. But subsequent runs should be fast.
:::

```{r}
#| label: test-keras
keras3::to_categorical(0)
```

If this prints a matrix without errors, your Keras installation is working correctly.