---
title: "Calculate and evaluate landscape metrics"
vignette: >
  %\VignetteIndexEntry{Calculate and evaluate landscape metrics}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
    cache: true
---

This guide shows how to calculate landscape metrics from `landscape` objects,
rank metrics by how well they distinguish known spatial patterns, and visualize
their distributions.

## Setup

Create a small set of labelled landscapes for the examples in this guide.

```{r}
#| label: setup
#| cache: false
library(patternscaper)

set.seed(123456)

# Generate sample landscapes
sample_landscapes <- create_landscapes(
  n = 20,
  patterns = c("random", "sharp", "labyrinth"),
  width = 50,
  height = 50
)
```

## Calculate landscape metrics

`calculate_metrics()` uses the
[`landscapemetrics` package](https://r-spatialecology.github.io/landscapemetrics/)
and calculates metrics at the landscape or class level:

- *Landscape-level* metrics summarize the complete landscape. They
  produce one value per landscape and metric.
- *Class-level* metrics summarize each land-cover class separately. They produce
  one value per landscape, metric, and land-cover class that is present.

Use landscape-level metrics to describe the complete landscape. Class-level
metrics can help distinguish patterns that differ mainly in which land-cover
class forms the pattern, such as spots and gaps.

Use `landscapemetrics::list_lsm()` to look up the available metric
abbreviations, names, and types at either level:

```{r}
#| label: list-metrics
# List available landscape-level metrics
landscapemetrics::list_lsm(level = "landscape") |>
  dplyr::select(metric, name, type)
```

By default (`metrics = NULL`), `calculate_metrics()` calculates every available
metric at the requested level:

```{r}
#| label: calculate-landscape-metrics
#| message: false
# Calculate all landscape-level metrics
landscape_metrics <- calculate_metrics(
  sample_landscapes,
  level = "landscape"
)
```

To calculate only specific metrics, you can pass their abbreviations to `metrics`.
This works at either level.

```{r}
#| label: calculate-selected-metrics
#| eval: false
# Calculate selected landscape-level metrics
landscape_specific_metrics <- calculate_metrics(
  sample_landscapes,
  level = "landscape",
  metrics = c("ai", "area_cv", "cai_cv")
)
```

In the resulting table, each row contains one metric result. The main columns identify the landscape, pattern, metric, calculated value, and any warning from `landscapemetrics`:

```{r}
#| label: inspect-landscape-metrics
landscape_metrics |>
  dplyr::select(
    landscape_id,
    landscape_name,
    pattern,
    metric,
    value,
    warnings
  )
```

In class-level results below, the `class` column stores the land-cover code. `metric_name` contains
the original metric abbreviation, while `metric` appends the class code to
give each result a unique identifier:

```{r}
#| label: calculate-class-metrics
#| message: false
# Calculate all class-level metrics
class_metrics <- calculate_metrics(
  sample_landscapes,
  level = "class"
)

class_metrics |>
  dplyr::select(
    landscape_id,
    pattern,
    class,
    metric_name,
    metric,
    value
  )
```

If a land-cover class is absent, its class-level metrics have no result row.
These incomplete metrics cannot be used directly as model predictors because
they do not provide one value per landscape.

:::{.callout-note}
## Comparable landscape geometry

Many landscape metrics change with raster extent, resolution, or shape. Use
comparable geometry where possible, or choose metrics that are robust to the
differences in your data. Metric models warn about substantial differences in
extent, resolution, or aspect ratio, but not about irregular shapes represented
by missing cells. See [Matching training and application
data](matching-training-application.html) for details.
:::

## Find informative metrics

Use `evaluate_metrics()` to rank metrics by how well they distinguish labelled
pattern classes in the supplied landscapes.

Three ranking methods are available:

- `kruskal_effsize` (default) uses ranks and is robust to outliers. It does not
  describe the size of differences on the original metric scale.
- `fisher_score` compares between- and within-pattern variance. It assumes
  approximately normally distributed values within patterns and is sensitive
  to outliers.
- `mean_groups` compares pattern means with the overall mean. It does
  not account for variation within patterns.

By default, `evaluate_metrics()` excludes incomplete and zero-variation
metrics because they cannot be used in model training.

After ranking, metrics are selected in rank order. A metric is skipped when its
absolute Pearson correlation with an already selected metric exceeds the
`correlation_threshold` (default 0.7). Correlations are calculated across all
landscapes and pattern classes. Setting `correlation_threshold` to 1 disables
correlation filtering.
If too few metrics pass the correlation filter, the highest-ranked
correlated metrics are added to reach `metrics_number`. They are marked
`selected_correlation_fill` and placed at the end of the selected metrics. Set
`fill_correlated = FALSE` to return only uncorrelated metrics (which then may be fewer than
requested).

The example below produces expected warnings about excluded metrics and
correlated metrics being added. They explain how the selection changed; they do
not indicate a failure.

```{r}
#| label: evaluate-metrics
# Evaluate metrics using the default Kruskal-Wallis effect-size method
metric_evaluation <- evaluate_metrics(
  metrics = landscape_metrics,
  method = "kruskal_effsize",
  metrics_number = 10,
  correlation_threshold = 0.7
)

metric_evaluation
```

The selected metric names are stored in `$selected`.

```{r}
#| label: view-selected-metrics
metric_evaluation$selected
```

The `$ranking` table contains the score and selection outcome for every input
metric:

```{r}
#| label: inspect-metric-ranking
metric_evaluation$ranking
```

The `name` column gives the full metric name including class information for class-level metrics.

The `outcome` column records whether a metric was selected, removed by the
correlation filter, or excluded before ranking. Excluded metrics have no score
or rank:

```{r}
#| label: summarize-ranking-outcomes
dplyr::count(metric_evaluation$ranking, outcome)
```

You can check which metrics were removed by the correlation filter and what they were
correlated with:

```{r}
#| label: inspect-correlated-metrics
metric_evaluation$ranking |>
  dplyr::filter(outcome == "dropped_correlated") |>
  dplyr::select(metric, score, correlated_with)
```

## Visualize metric distributions

Use `plot_metrics()` to compare the ten selected metrics. Panel labels use
metric abbreviations by default. Set `metric_labels = "name"` to show full names
from `landscapemetrics::list_lsm()`. `jitter_seed` fixes the jittered point
positions so the figure is reproduced exactly:

```{r}
#| label: plot-metrics
#| fig-alt: "Boxplots of ten selected landscape metrics with full metric names"
#| fig-cap: "The selected landscape metrics labelled with their full names."
#| fig-width: 8
#| fig-height: 6
plot_metrics(
  landscape_metrics,
  selected_metrics = metric_evaluation,
  metric_labels = "name",
  jitter_seed = 42
)
```

Long names wrap automatically. Set `label_wrap_width` if the labels look too
wide or too narrow for the figure size:

```{r}
#| label: customize-label-wrapping
#| eval: false
plot_metrics(
  landscape_metrics,
  selected_metrics = metric_evaluation,
  metric_labels = "name",
  label_wrap_width = 15
)
```

## Next steps

Continue with [Classify landscapes with landscape metrics](classify-metrics.html)
to train and evaluate a classifier and apply it to new landscapes.
