---
title: "Create landscapes with different spatial patterns"
vignette: >
  %\VignetteIndexEntry{landscape-generation}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

Create artificial landscapes with known spatial patterns. The [metric-based classification guide](classify-metrics.html) and [pixel-based classification guide](classify-pixels.html) show how to use the created landscapes for training and testing classifiers.

```{r}
#| label: setup
library(patternscaper)
# Set seed for reproducibility
set.seed(123456)
```

## Overview

Landscape generation uses three kinds of functions:

- `create_landscape()` generates one landscape
- `create_landscapes()` generates a batch of landscapes
- `pattern_*()` constructors set pattern-specific parameters

Pass a matching `pattern_*()` constructor to `params` or `params_list`:

```{r}
#| label: api-shape
#| fig-width: 4
#| fig-height: 3
#| fig-alt: "A single spots landscape with 8 large spots"

big_spots <- create_landscape(
  "spots",
  params = pattern_spots(n_spots = 8, spot_radius = 12)
)

plot_landscapes(big_spots, show_legend = FALSE)
```

## The patterns

The package provides 11 spatial patterns. Each `pattern_*` help page documents the parameters you can set for the corresponding landscape.

### Control patterns

**random**, **bare**, and **dense** place vegetation at random. They differ only in `veg_prob`, the proportion of vegetated cells: `bare` defaults to 0.1, `random` to 0.5, and `dense` to 0.9. These patterns provide control landscapes without a specific spatial structure.

```{r}
#| label: control-patterns
#| fig-width: 8
#| fig-height: 3
#| fig-alt: "Three landscapes showing control patterns: bare, random, and dense"

bare <- create_landscape("bare", name = "bare (veg_prob 0.1)")
random <- create_landscape("random", name = "random (veg_prob 0.5)")
dense <- create_landscape("dense", name = "dense (veg_prob 0.9)")

plot_landscapes(
  list(bare, random, dense),
  titles = "name",
  show_legend = FALSE
)
```

### Ecotone patterns

Ecotone patterns have a vegetated zone and a bare zone separated by a transition. `boundary_position` places that transition in all five.

#### sharp

An abrupt boundary, with `boundary_position` as its only parameter.

```{r}
#| label: pattern-sharp
#| fig-width: 8
#| fig-height: 3
#| fig-alt: "Two sharp-boundary landscapes: default and a low boundary"

plot_landscapes(
  list(
    create_landscape("sharp", name = "default"),
    create_landscape(
      "sharp",
      name = "boundary_position = 0.3",
      params = pattern_sharp(boundary_position = 0.3)
    )
  ),
  titles = "name",
  show_legend = FALSE
)
```

#### diffuse

A gradual transition, where the chance of a cell being vegetated falls off with distance from the boundary. `steepness` runs from 0 to 1 and sets how wide that transition zone is: **higher** values spread it over more of the landscape.

```{r}
#| label: pattern-diffuse
#| fig-width: 8
#| fig-height: 3
#| fig-alt: "Three diffuse-boundary landscapes with increasingly wide transition zones"

plot_landscapes(
  list(
    create_landscape(
      "diffuse",
      name = "steepness = 0.1 (narrow)",
      params = pattern_diffuse(steepness = 0.1)
    ),
    create_landscape("diffuse", name = "default (0.5)"),
    create_landscape(
      "diffuse",
      name = "steepness = 1 (broad)",
      params = pattern_diffuse(steepness = 1)
    )
  ),
  titles = "name",
  show_legend = FALSE
)
```

#### fingers

Curving, finger-like extensions of vegetation into the bare zone. The parameter `sine_length_mean` controls the average wavelength of the curves (larger values result in longer, more widely spaced bends). The parameter `sine_height_mean` controls the average finger amplitude (larger values result in more pronounced bends). The corresponding `*_sd` parameters control the variation between fingers.

```{r}
#| label: pattern-fingers
#| fig-width: 8
#| fig-height: 3
#| fig-alt: "Three finger landscapes: default, long pronounced bends, and short pronounced bends"

plot_landscapes(
  list(
    create_landscape("fingers", name = "default"),
    create_landscape(
      "fingers",
      name = "long, pronounced bends",
      params = pattern_fingers(
        sine_length_mean = 40,
        sine_height_mean = 12
      )
    ),
    create_landscape(
      "fingers",
      name = "short, pronounced bends",
      params = pattern_fingers(
        sine_length_mean = 10,
        sine_height_mean = 15
      )
    )
  ),
  titles = "name",
  show_legend = FALSE
)
```

#### clustered

Vegetation clusters scattered above the boundary. `n_clusters` and `cluster_radius` set how many and how big, `cluster_zone` how far into the bare zone they reach, and `elongation_x` / `elongation_y` stretch them along one axis.

```{r}
#| label: pattern-clustered
#| fig-width: 8
#| fig-height: 3
#| fig-alt: "Three clustered landscapes: default, few large clusters, and many small elongated clusters"

plot_landscapes(
  list(
    create_landscape("clustered", name = "default"),
    create_landscape(
      "clustered",
      name = "few, large",
      params = pattern_clustered(n_clusters = 5, cluster_radius = 10)
    ),
    create_landscape(
      "clustered",
      name = "many, small, elongated",
      params = pattern_clustered(
        n_clusters = 20,
        cluster_radius = 3,
        elongation_x = 3
      )
    )
  ),
  titles = "name",
  show_legend = FALSE
)
```

#### bands

Sinusoidal vegetation bands running parallel to the boundary. `band_thickness` and `band_spacing` set their geometry, `amplitude` sets the height of their waves, and `frequency` sets how often they wave.

```{r}
#| label: pattern-bands
#| fig-width: 8
#| fig-height: 3
#| fig-alt: "Three banded landscapes: default, thick widely spaced bands, and bands with high, frequent waves"

plot_landscapes(
  list(
    create_landscape("bands", name = "default"),
    create_landscape(
      "bands",
      name = "thick, widely spaced",
      params = pattern_bands(band_thickness = 6, band_spacing = 20)
    ),
    create_landscape(
      "bands",
      name = "high, frequent waves",
      params = pattern_bands(amplitude = 15, frequency = 0.15)
    )
  ),
  titles = "name",
  show_legend = FALSE
)
```

### Patch patterns

Self-organized patterns, without a boundary between two zones.

#### spots

Circular vegetation patches on bare ground. `n_spots` and `spot_radius` set the count and size, `spot_radius_sd` varies the size between spots, `radius_noise_fraction` roughens their edges, and `regular_spots = TRUE` places them on a grid instead of at random.

```{r}
#| label: pattern-spots
#| fig-width: 8
#| fig-height: 3
#| fig-alt: "Three spot landscapes: default, fewer larger variable spots, and a regular grid"

plot_landscapes(
  list(
    create_landscape("spots", name = "default"),
    create_landscape(
      "spots",
      name = "fewer, larger, variable",
      params = pattern_spots(
        n_spots = 3,
        spot_radius = 12,
        spot_radius_sd = 3
      )
    ),
    create_landscape(
      "spots",
      name = "regular grid",
      params = pattern_spots(
        n_spots = 9,
        spot_radius = 8,
        regular_spots = TRUE
      )
    )
  ),
  titles = "name",
  show_legend = FALSE
)
```

#### gaps

The inverse of spots: circular bare gaps in otherwise vegetated ground. It takes parameters analogous to those for spots, with names such as `n_gaps` and `gap_radius`.

```{r}
#| label: pattern-gaps
#| fig-width: 8
#| fig-height: 3
#| fig-alt: "Three gap landscapes: default, few large gaps, and many gaps with ragged edges"

plot_landscapes(
  list(
    create_landscape("gaps", name = "default"),
    create_landscape(
      "gaps",
      name = "few, large",
      params = pattern_gaps(n_gaps = 5, gap_radius = 15)
    ),
    create_landscape(
      "gaps",
      name = "ragged edges",
      params = pattern_gaps(
        n_gaps = 12,
        gap_radius = 8,
        radius_noise_fraction = 0.4
      )
    )
  ),
  titles = "name",
  show_legend = FALSE
)
```

#### labyrinth

Maze-like bands of vegetation. `frequency` sets the scale of the maze and `octaves` how much fine detail is layered on top; `veg_threshold` shifts the overall vegetated proportion.

```{r}
#| label: pattern-labyrinth
#| fig-width: 8
#| fig-height: 3
#| fig-alt: "Three labyrinth landscapes: default, a finer maze, and a labyrinth with greater vegetation cover"

plot_landscapes(
  list(
    create_landscape("labyrinth", name = "default"),
    create_landscape(
      "labyrinth",
      name = "finer, more detailed",
      params = pattern_labyrinth(frequency = 6, octaves = 3)
    ),
    create_landscape(
      "labyrinth",
      name = "greater vegetation cover",
      params = pattern_labyrinth(veg_threshold = 0.35)
    )
  ),
  titles = "name",
  show_legend = FALSE
)
```

## Creating training landscapes

The quickest way to generate training data is `create_landscapes()`. It generates multiple landscapes, distributes them as evenly as possible across the selected patterns, and varies their parameters between landscapes.

```{r}
#| label: create-training-landscapes
#| fig-width: 8
#| fig-height: 6
#| fig-alt: "Grid showing 20 randomly generated landscapes with different spatial patterns"

# Create 20 landscapes distributed across all pattern types
landscapes <- create_landscapes(n = 20)
```

The result is a list of [`landscape` objects](#the-landscape-object).

Plot them with `plot_landscapes()`:

```{r}
#| label: plot-training-landscapes
#| fig-width: 8
#| fig-height: 6
#| fig-alt: "Grid of 20 landscapes distributed across the 11 available spatial patterns"
# Plot all landscapes
plot_landscapes(landscapes)
```

By default, landscapes are distributed as evenly as possible across the selected pattern types. When the number of landscapes is not divisible by the number of patterns, some patterns occur once more than others.

Inspect the number of generated patterns with `table()`:

```{r}
#| label: check-pattern-distribution

# Check how many landscapes of each type were generated
table(purrr::map_chr(landscapes, \(x) x$pattern))
```

### Selecting specific patterns

Specific patterns can be selected to create only a subset of patterns using the `patterns` argument. For example, to create landscapes with only labyrinth, spots, and clustered patterns:

```{r}
#| label: specific-patterns
#| fig-width: 8
#| fig-height: 6
#| fig-alt: "Grid showing landscapes with only labyrinth, spots, and clustered patterns"

# Generate only specific patterns
landscapes <- create_landscapes(
  n = 12,
  patterns = c("labyrinth", "spots", "clustered")
)

plot_landscapes(landscapes)
```

### Landscape size

To modify the size (number of pixels in x- and y-direction) of all generated landscapes, use the `width` and `height` arguments. By default, landscapes are created with a size of 100 x 100 pixels.

```{r}
#| label: custom-size
#| fig-width: 8
#| fig-height: 2
#| fig-alt: "Three non-square landscapes of 50 x 20 pixels"
non_square <- create_landscapes(
  n = 3,
  width = 50,
  height = 20
)
# Plot landscapes with custom size
plot_landscapes(non_square)
```

### Landscape rotation

By default, each rotatable pattern is rotated by a whole-degree angle sampled between 0 and 360. This shows classifiers similar landscapes in different orientations during training and reduces reliance on orientation for the classification.

Set `rotation = 0` to disable rotation, or supply a length-2 range from which angles are sampled. This example uses angles between 45 and 90 degrees:

```{r}
#| label: rotation-example
#| message: false
#| fig-width: 8
#| fig-height: 3
#| fig-alt: "Three unrotated clustered, sharp, and banded landscapes followed by three clustered landscapes rotated between 45 and 90 degrees"

no_rotation <- create_landscapes(
  n = 3,
  patterns = c("clustered", "sharp", "bands"),
  rotation = 0
)

defined_angles <- create_landscapes(
  n = 3,
  patterns = "clustered",
  rotation = c(45, 90)
)

plot_landscapes(c(no_rotation, defined_angles))
```

Only "sharp", "diffuse", "fingers", "clustered" and "bands" are rotated. The remaining patterns ignore `rotation`, since a batch can mix rotatable and non-rotatable patterns.

### Pattern-specific parameters

The `params_list` argument sets parameters by pattern name using the same `pattern_*()` constructors. Unlike `create_landscape()`, a batch parameter can be either a length-2 range to sample the parameter from or a single value used for all landscapes of that pattern:

```{r}
#| label: custom-parameters
#| message: false
#| fig-width: 8
#| fig-height: 6
#| fig-alt: "Grid showing landscapes with many large spots of varying size"

bigger_spots <- create_landscapes(
  n = 12,
  patterns = "spots",
  params_list = list(
    spots = pattern_spots(
      n_spots = c(10, 20), # range for each landscape
      spot_radius = c(8, 12), # range for each landscape
      spot_radius_sd = 3, # single value for all landscapes
      regular_spots = FALSE # single value for all landscapes
    )
  )
)

plot_landscapes(bigger_spots)
```

A single value fixes a parameter for every landscape in the batch; a length-2 vector is a range. Above, `n_spots` and `spot_radius` vary between landscapes, while `spot_radius_sd` and `regular_spots` are the same throughout. Patterns left out of `params_list` fall back to their default sampling ranges, which are listed on each constructor's help page.

## The landscape object

In `patternscaper`, landscapes are lists containing raster data and metadata. All landscape generation functions return landscapes in this format and all other functions expect landscapes to be provided in this format.

To convert other raster data to this format, see the [importing landscapes vignette](importing-landscapes.html).

Each landscape is stored as a list with the following components:

- `data`: landscape raster as a `SpatRaster`
- `pattern`: pattern label, such as `"spots"` or `"clustered"`
- `name`: user-defined name
- `params`: parameters used to generate the landscape

```{r}
#| label: landscape-object-structure

landscape_example <- create_landscape("spots", name = "Example")
str(landscape_example)
```

### Modifying landscape metadata

Users can modify landscape metadata using the helper functions `set_landscape_pattern()` and `set_landscape_name()`. For example, to change the pattern label or name of existing landscapes:

```{r}
#| label: modify-metadata

landscapes_example <- set_landscape_name(
  landscape_example,
  "New Landscape Name"
)
landscapes_example <- set_landscape_pattern(
  landscapes_example,
  "ecotone"
)
landscapes_example
```

## Next steps

- [Calculate landscape metrics](landscape-metrics.html): Extract landscape features and identify informative metrics
- [Classify with landscape metrics](classify-metrics.html): Train classifiers on landscape metrics
- [Classify with pixels](classify-pixels.html): Train Keras classifiers on raster-cell values
