---
title: "Generate landscapes with different spatial patterns"
vignette: >
  %\VignetteIndexEntry{landscape-generation}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

This vignette shows how to generate artificial landscapes with different spatial patterns. These landscapes are designed for training and testing spatial pattern classifiers. For details on how to use the generated landscapes for training classifiers, see the [classify-metrics](classify-metrics.html) and [classify-pixels](classify-pixels.html) vignettes.

```{r}
#| label: setup

library(spatPatClassifyR)
# Set seed for reproducibility
set.seed(123456)
```

## Overview

Landscape generation uses three kinds of function:

- `create_landscapes()` generates a batch of landscapes for training
- `create_landscape()` generates a single landscape
- `pattern_random()`, `pattern_spots()`, ... set the parameters of one pattern

The first two are the entry points. Each pattern has its own parameters, and those are set through the pattern's own constructor, which is passed to `params`:

```{r}
#| label: api-shape
#| fig-width: 4
#| fig-height: 3
#| fig-alt: "A single spots landscape with 8 large spots"

big_spots <- create_landscape(
  "spots",
  params = pattern_spots(n_spots = 8, spot_radius = 12)
)

plot_landscapes(list(big_spots), show_legend = FALSE)
```

Typing `pattern_` and pressing <kbd>Tab</kbd> lists all 11 patterns, and each constructor's arguments are that pattern's parameters -- so autocomplete works, and `?pattern_spots` documents them.

## The patterns

Currently there are 11 spatial patterns. This section shows what each one looks like and which parameters shape it.

### Control patterns

**random**, **bare** and **dense** all place vegetation at random. They differ only in `veg_prop`, the proportion of vegetated cells: `bare` defaults to 0.1, `random` to 0.5 and `dense` to 0.9. They serve as controls -- landscapes with no spatial structure at all.

```{r}
#| label: control-patterns
#| fig-width: 8
#| fig-height: 3
#| fig-alt: "Three landscapes showing control patterns: bare, random, and dense"

bare <- create_landscape("bare", name = "bare (veg_prop 0.1)")
random <- create_landscape("random", name = "random (veg_prop 0.5)")
dense <- create_landscape("dense", name = "dense (veg_prop 0.9)")

plot_landscapes(
  list(bare, random, dense),
  titles = "name",
  show_legend = FALSE
)
```

See `?pattern_random`, `?pattern_bare` and `?pattern_dense`.

### Ecotone patterns

Ecotone patterns have a vegetated zone and a bare zone separated by a transition. `boundary_position` places that transition in all five.

#### sharp

An abrupt boundary, with `boundary_position` as its only parameter.

```{r}
#| label: pattern-sharp
#| fig-width: 8
#| fig-height: 3
#| fig-alt: "Two sharp-boundary landscapes: default and a low boundary"

plot_landscapes(
  list(
    create_landscape("sharp", name = "default"),
    create_landscape(
      "sharp",
      name = "boundary_position = 0.3",
      params = pattern_sharp(boundary_position = 0.3)
    )
  ),
  titles = "name",
  show_legend = FALSE
)
```

See `?pattern_sharp`.

#### diffuse

A gradual transition, where the chance of a cell being vegetated falls off with distance from the boundary. `steepness` runs from 0 to 1 and sets how wide that transition zone is: **higher** values spread it over more of the landscape.

```{r}
#| label: pattern-diffuse
#| fig-width: 8
#| fig-height: 3
#| fig-alt: "Three diffuse-boundary landscapes with increasingly wide transition zones"

plot_landscapes(
  list(
    create_landscape(
      "diffuse",
      name = "steepness = 0.1 (narrow)",
      params = pattern_diffuse(steepness = 0.1)
    ),
    create_landscape("diffuse", name = "default (0.5)"),
    create_landscape(
      "diffuse",
      name = "steepness = 1 (broad)",
      params = pattern_diffuse(steepness = 1)
    )
  ),
  titles = "name",
  show_legend = FALSE
)
```

See `?pattern_diffuse`.

#### fingers

Curvy, finger-like extensions of vegetation into the bare zone. Their width is set by `sine_length_mean` and how far they reach by `sine_height_mean`, each with a matching `_sd` for variation between fingers.

```{r}
#| label: pattern-fingers
#| fig-width: 8
#| fig-height: 3
#| fig-alt: "Three finger landscapes: default, broad fingers, and narrow tall fingers"

plot_landscapes(
  list(
    create_landscape("fingers", name = "default"),
    create_landscape(
      "fingers",
      name = "broad, far-reaching",
      params = pattern_fingers(sine_length_mean = 40, sine_height_mean = 12)
    ),
    create_landscape(
      "fingers",
      name = "narrow, tall",
      params = pattern_fingers(sine_length_mean = 10, sine_height_mean = 15)
    )
  ),
  titles = "name",
  show_legend = FALSE
)
```

See `?pattern_fingers`.

#### clustered

Vegetation clusters scattered above the boundary. `n_clusters` and `cluster_radius` set how many and how big, `scatter_zone_prop` how far into the bare zone they reach, and `elongation_x` / `elongation_y` stretch them along one axis.

```{r}
#| label: pattern-clustered
#| fig-width: 8
#| fig-height: 3
#| fig-alt: "Three clustered landscapes: default, few large clusters, and many small elongated clusters"

plot_landscapes(
  list(
    create_landscape("clustered", name = "default"),
    create_landscape(
      "clustered",
      name = "few, large",
      params = pattern_clustered(n_clusters = 5, cluster_radius = 10)
    ),
    create_landscape(
      "clustered",
      name = "many, small, elongated",
      params = pattern_clustered(
        n_clusters = 20,
        cluster_radius = 3,
        elongation_x = 3
      )
    )
  ),
  titles = "name",
  show_legend = FALSE
)
```

See `?pattern_clustered`.

#### bands

Sinusoidal vegetation bands running parallel to the boundary. `band_thickness` and `band_spacing` set their geometry, `amplitude` and `frequency` how strongly they wave.

```{r}
#| label: pattern-bands
#| fig-width: 8
#| fig-height: 3
#| fig-alt: "Three banded landscapes: default, thick widely spaced bands, and strongly waved bands"

plot_landscapes(
  list(
    create_landscape("bands", name = "default"),
    create_landscape(
      "bands",
      name = "thick, widely spaced",
      params = pattern_bands(band_thickness = 6, band_spacing = 20)
    ),
    create_landscape(
      "bands",
      name = "strongly waved",
      params = pattern_bands(amplitude = 15, frequency = 0.15)
    )
  ),
  titles = "name",
  show_legend = FALSE
)
```

See `?pattern_bands`.

### Patch patterns

Self-organized patterns, without a boundary between two zones.

#### spots

Circular vegetation patches on bare ground. `n_spots` and `spot_radius` set the count and size, `spot_radius_sd` varies the size between spots, `radius_noise_fraction` roughens their edges, and `regular_spots = TRUE` places them on a grid instead of at random.

```{r}
#| label: pattern-spots
#| fig-width: 8
#| fig-height: 3
#| fig-alt: "Three spot landscapes: default, fewer larger variable spots, and a regular grid"

plot_landscapes(
  list(
    create_landscape("spots", name = "default"),
    create_landscape(
      "spots",
      name = "fewer, larger, variable",
      params = pattern_spots(
        n_spots = 8,
        spot_radius = 12,
        spot_radius_sd = 3
      )
    ),
    create_landscape(
      "spots",
      name = "regular grid",
      params = pattern_spots(
        n_spots = 9,
        spot_radius = 8,
        regular_spots = TRUE
      )
    )
  ),
  titles = "name",
  show_legend = FALSE
)
```

See `?pattern_spots`.

#### gaps

The inverse of spots: circular bare gaps in otherwise vegetated ground. It takes the same parameters, except that inverting is what makes it `gaps`, so that is not adjustable.

```{r}
#| label: pattern-gaps
#| fig-width: 8
#| fig-height: 3
#| fig-alt: "Three gap landscapes: default, few large gaps, and many gaps with ragged edges"

plot_landscapes(
  list(
    create_landscape("gaps", name = "default"),
    create_landscape(
      "gaps",
      name = "few, large",
      params = pattern_gaps(n_gaps = 5, gap_radius = 15)
    ),
    create_landscape(
      "gaps",
      name = "ragged edges",
      params = pattern_gaps(
        n_spots = 12,
        spot_radius = 8,
        radius_noise_fraction = 0.4
      )
    )
  ),
  titles = "name",
  show_legend = FALSE
)
```

See `?pattern_gaps`.

#### labyrinth

Maze-like bands of vegetation. `frequency` sets the scale of the maze and `octaves` how much fine detail is layered on top; `veg_threshold` shifts the overall vegetated proportion.

```{r}
#| label: pattern-labyrinth
#| fig-width: 8
#| fig-height: 3
#| fig-alt: "Three labyrinth landscapes: default, a finer maze, and a sparser maze"

plot_landscapes(
  list(
    create_landscape("labyrinth", name = "default"),
    create_landscape(
      "labyrinth",
      name = "finer, more detailed",
      params = pattern_labyrinth(frequency = 6, octaves = 3)
    ),
    create_landscape(
      "labyrinth",
      name = "veg_threshold = 0.35",
      params = pattern_labyrinth(veg_threshold = 0.35)
    )
  ),
  titles = "name",
  show_legend = FALSE
)
```

See `?pattern_labyrinth`.

## Creating training landscapes

The quickest way to generate training data is `create_landscapes()`. It generates multiple landscapes balanced across all available patterns, varying their parameters between landscapes.

```{r}
#| label: create-training-landscapes
#| fig-width: 8
#| fig-height: 6
#| fig-alt: "Grid showing 20 randomly generated landscapes with different spatial patterns"

# Create 20 landscapes balanced across all pattern types
landscapes <- create_landscapes(n = 20)
```

The `landscapes` are returned as a list of `landscape` objects (see [below](landscape-generation.qmd#the-landscape-object) for more info on the data structure).

They can be visualized using the `plot_landscapes()` function:

```{r}
#| label: plot-training-landscapes
#| fig-width: 8
#| fig-height: 6
# Plot all landscapes
plot_landscapes(landscapes)
```

By default, landscapes are created in a balanced way, so that each pattern type is represented equally in the generated set.

```{r}
#| label: check-pattern-distribution

# Check how many landscapes of each type were generated
table(purrr::map_chr(landscapes, ~ .x$pattern))
```

### Selecting specific patterns

Specific patterns can be selected to create only a subset of patterns using the `patterns` argument. For example, to create landscapes with only labyrinth, spots, and clustered patterns:

```{r}
#| label: specific-patterns
#| fig-width: 8
#| fig-height: 6
#| fig-alt: "Grid showing landscapes with only labyrinth, spots, and clustered patterns"

# Generate only specific patterns
landscapes <- create_landscapes(
  n = 12,
  patterns = c("labyrinth", "spots", "clustered")
)

plot_landscapes(landscapes)
```

### Landscape size

To modify the size (number of pixels in x- and y-direction) of all generated landscapes, use the `width` and `height` arguments. By default, landscapes are created with a size of 100x100 pixels.

```{r}
#| label: custom-size
#| fig-width: 8
#| fig-height: 2
#| fig-alt: "Three non-square landscapes of 50x20 pixels"
non_square <- create_landscapes(
  n = 3,
  width = 50,
  height = 20
)
# Plot landscapes with custom size
plot_landscapes(non_square)
```

### Landscape rotation

By default, landscapes are rotated by angles randomly chosen between 0 and 360 degrees.
This ensures that classifiers trained on these landscapes are invariant to landscape orientation.

You can disable random rotation by setting the `rotation` parameter to `0`.
This parameter can also be used to directly control the range of selected rotation angles. For example, to select only angles between 0 and 90 degrees:

```{r}
#| label: rotation-example
#| message: false
#| fig-width: 8
#| fig-height: 3
#| fig-alt: "Landscapes showing the same pattern rotated at different angles"

no_rotation <- create_landscapes(
  n = 3,
  patterns = c("clustered", "sharp", "bands"),
  rotation = 0
)

defined_angles <- create_landscapes(
  n = 3,
  patterns = "clustered",
  rotation = c(45, 90)
)

plot_landscapes(c(no_rotation, defined_angles))
```

Only "sharp", "diffuse", "fingers", "clustered" and "bands" are rotated. The remaining patterns ignore `rotation`, since a batch can mix rotatable and non-rotatable patterns.

### Pattern-specific parameters

The `params_list` argument sets parameters per pattern, keyed by pattern name, using the same `pattern_*()` constructors. The difference from `create_landscape()` is that here a parameter may be given as a **range** -- a length-2 vector -- which is sampled once per landscape:

```{r}
#| label: custom-parameters
#| message: false
#| fig-width: 8
#| fig-height: 6
#| fig-alt: "Grid showing landscapes with many large spots of varying size"

bigger_spots <- create_landscapes(
  n = 12,
  patterns = "spots",
  params_list = list(
    spots = pattern_spots(
      n_spots = c(10, 20),
      spot_radius = c(8, 12),
      spot_radius_sd = 3,
      regular_spots = FALSE
    )
  )
)

plot_landscapes(bigger_spots)
```

A single value fixes a parameter for every landscape in the batch; a length-2 vector is a range. Above, `n_spots` and `spot_radius` vary between landscapes, while `spot_radius_sd` and `regular_spots` are the same throughout. Patterns left out of `params_list` fall back to their default sampling ranges, which are listed on each constructor's help page.

## The landscape object

In `spatPatClassifyR`, landscapes are stored as lists containing the landscape data and metadata. All landscape generation functions return landscapes in this format and all other functions expect landscapes to be provided in this format.

To learn how to transform your own raster data into this format, see [vignette on importing landscapes](importing-landscapes.html).

Each landscape is stored as a list with the following components:

- data: The landscape object as a `SpatRaster` class
- pattern: Pattern type (e.g., "spots", "clustered")
- name: User-defined name
- params: List of parameters used to generate the landscape

```{r}
#| label: landscape-object-structure

landscape_example <- create_landscape("spots", name = "Example")
str(landscape_example)
```

### Modifying landscape metadata

Users can modify landscape metadata using the helper functions `set_landscape_pattern()` and `set_landscape_name()`. For example, to change the pattern label or name of existing landscapes:

```{r}
#| label: modify-metadata

landscapes_example <- set_landscape_name(
  landscape_example,
  "New Landscape Name"
)
landscapes_example <- set_landscape_pattern(
  landscapes_example,
  "ecotone"
)
landscapes_example
```

## Next Steps

- See `vignette("landscape-metrics")` for extracting landscape features and finding the most informative metrics
- See `vignette("classify-metrics")` for training spatial pattern classifiers on the landscape metrics
- See `vignette("classify-pixels")` for training spatial pattern classifiers on the raw landscape data using Keras
