---
title: "Calculate and evaluate landscape metrics"
vignette: >
  %\VignetteIndexEntry{Calculate and evaluate landscape metrics}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
    cache: true
---

```{r}
#| label: setup
library(spatPatClassifyR)

set.seed(123456)

# Generate sample landscapes
sample_landscapes <- create_landscapes(
  n = 20,
  patterns = c("random", "sharp", "labyrinth"),
  width = 50,
  height = 50
)
```

This vignette shows how to calculate landscape metrics from `landscape` objects and how to find metrics that are informative for distinguishing different spatial patterns.

## Calculating Landscape Metrics

Landscape metrics can be calculated using the `calculate_metrics()` function. By default, this function computes all landscape metrics available in the [`landscapemetrics` package](https://r-spatialecology.github.io/landscapemetrics/). Users can also specify a subset of metrics to calculate by providing a vector of metric names. To find available metric names, use:

```{r}
# Select either the landscape-level or class-level metrics
landscapemetrics::list_lsm(level = "landscape")
```

Metrics can be calculated on the landscape level (one metric for the entire landscape) or on the class level (one metric per land cover class in the landscape).

```{r}
# Calculate all landscape-level metrics for the sample landscapes
landscape_metrics <- calculate_metrics(
  sample_landscapes,
  level = "landscape"
)
```

To calculate selected metrics, provide a vector of valid metric names to the `metrics` argument. Find available metric names using `landscapemetrics::list_lsm()` as shown above.

```{r}
# Calculate all class-level metrics for the sample landscapes
class_metrics <- calculate_metrics(
  sample_landscapes,
  level = "class",
  metrics = c("ai", "area_cv", "cai_cv") # Define a subset of metrics to calculate
)
```

## Finding Informative Metrics

The package includes different methods to evaluate the calculated landscape metrics and find those that are most informative for distinguishing different spatial patterns. The general function to evaluate landscape metrics is `evaluate_metrics()`. It supports different methods for metric evaluation:

- `coeffvar_all`: Coefficient of Variation (CV = SD/mean). Ranks metrics by their relative variability across landscapes.
- `lin_mod_r2`: Fits value ~ pattern for each metric and ranks by R².
- `mean_groups`: Calculates relative differences between pattern-specific means and overall mean, then sums across patterns.
- `fisher_score`: Fisher Score (ratio of between-group to within-group variance). 
- `kruskal_effsize`: Kruskal-Wallis H test effect sizes. Non-parametric test for differences between groups.

By default, the function excludes metrics with high correlation (above 0.7). This can be adjusted using the `correlation_threshold` argument.

```{r}
# Evaluate landscape metrics using Fisher Score
metric_evaluation <- evaluate_metrics(
  landscape_metrics,
  method = "fisher_score",
  metrics_number = 10, # Select top 10 informative metrics
  correlation_threshold = 0.7
)

metric_evaluation
```

The selected metric names are in `$selected`, ready to pass on to
`plot_metrics()` or `train_metrics_model()`. You can also pass the whole object
to either function.

```{r}
metric_evaluation$selected
```

### Inspecting the ranking

Selection discards a lot of detail: which metrics scored well but lost out to a
correlated partner, which never made it into the ranking at all, and how large
the differences between the top metrics actually were. `$ranking` keeps all of
it, with one row for every metric you passed in:

```{r}
metric_evaluation$ranking
```

The `name` column spells out what each abbreviation stands for, so you do not
have to look them up separately. Many metrics summarise a value across all
patches and come as a `_cv`/`_mn`/`_sd` triple that `landscapemetrics` gives a
single name, so the statistic is added in brackets to keep them apart:
`area_mn` is `Patch area (mean)`. For class-level metrics the class is added
too, for example `Patch area (mean, class 1)`.

The `outcome` column records what happened to each candidate. Metrics with an
`excluded_` outcome never reached the ranking, so they have no score or rank:

```{r}
dplyr::count(metric_evaluation$ranking, outcome)
```

This is what to look at when a selection surprises you. For example, to see the
metrics that were dropped for being too similar to an already selected one, and
what they clashed with:

```{r}
metric_evaluation$ranking |>
  dplyr::filter(outcome == "dropped_correlated") |>
  dplyr::select(metric, score, correlated_with)
```

If fewer uncorrelated metrics exist than you asked for, the remaining slots are
filled with correlated ones. Those are marked `selected_correlation_fill` and
appear at the end of `$selected`, so you can always tell which of your selected
metrics are redundant.

To visually compare the top 10 metrics, use the `plot_metrics()` function:

```{r}
#| fig-alt: "Boxplots showing the distribution of the top 10 informative landscape metrics across different spatial patterns"
#| fig-cap: "Distribution of the top 10 informative landscape metrics across different spatial patterns. Each dot is a landscape, the boxplots show the distribution of metric values for each spatial pattern."
#| fig-width: 8
#| fig-height: 6
plot_metrics(
  landscape_metrics,
  selected_metrics = metric_evaluation
)
```

The panel labels are the metric abbreviations, which are compact but not very
readable unless you already know them. Set `metric_labels = "name"` to label the
panels with the full metric names from `landscapemetrics::list_lsm()` instead:

```{r}
#| fig-alt: "The same boxplots, with panels labelled by full metric names instead of abbreviations"
#| fig-cap: "The same metrics, labelled with their full names. Useful for plots that go into a talk or a paper, where the abbreviations would need explaining."
#| fig-width: 8
#| fig-height: 6
plot_metrics(
  landscape_metrics,
  selected_metrics = metric_evaluation,
  metric_labels = "name"
)
```

Long names are wrapped over several lines. The width is chosen from the number
of panel columns, which works for most selections, but it counts characters
rather than measuring the rendered text — so if your labels look cramped or too
sparse at your figure size, set `label_wrap_width` yourself:

```{r}
#| eval: false
plot_metrics(
  landscape_metrics,
  selected_metrics = metric_evaluation,
  metric_labels = "name",
  label_wrap_width = 15
)
```

For class-level metrics the class is appended to the name, for example
`Aggregation index (class 1)`, so the two classes stay distinguishable.