---
title: "Create artificial landscapes"
vignette: >
  %\VignetteIndexEntry{Create artificial landscapes}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

Generate single artificial landscapes or batches with known spatial patterns for training and testing classifiers. If you are new to the package, begin with [Get started](patternscaper.html) for an overview of the complete classification workflow.

```{r}
#| label: setup
library(patternscaper)
# Set seed for reproducibility
set.seed(123456)
```

## Overview

Landscape generation uses three kinds of functions:

- `create_landscape()` generates one landscape
- `create_landscapes()` generates a batch of landscapes
- `pattern_*()` constructors set pattern-specific parameters

Pass a matching `pattern_*()` constructor to `params` or `params_list`:

```{r}
#| label: api-shape
#| fig-width: 4
#| fig-height: 3
#| fig-alt: "A single spots landscape with 8 large spots"

big_spots <- create_landscape(
  "spots",
  params = pattern_spots(n_spots = 8, spot_radius = 12)
)

plot_landscapes(big_spots, show_legend = FALSE)
```

## Choose a pattern

The package provides 11 spatial patterns in three groups: control, ecotone, and patch patterns. See the [pattern gallery](pattern-gallery.html) to compare them and explore their parameters.

## Creating training landscapes

The quickest way to generate training data is `create_landscapes()`. It generates multiple landscapes, distributes them as evenly as possible across the selected patterns, and varies their parameters between landscapes.

The following example creates only 20 landscapes to keep the guide quick to run. For training the actual classifiers, the data set should be larger.

```{r}
#| label: create-training-landscapes

# Create 20 landscapes distributed across all pattern types
landscapes <- create_landscapes(n = 20)
```

The result is a list of [`landscape` objects](#landscape-objects).

Plot them with `plot_landscapes()`:

```{r}
#| label: plot-training-landscapes
#| fig-width: 8
#| fig-height: 6
#| fig-alt: "Grid of 20 landscapes distributed across the 11 available spatial patterns"
# Plot all landscapes
plot_landscapes(landscapes)
```

By default, landscapes are distributed as evenly as possible across the selected pattern types. When the number of landscapes is not divisible by the number of patterns, some patterns occur once more than others.

Inspect the number of generated patterns with `table()`:

```{r}
#| label: check-pattern-distribution

# Check how many landscapes of each type were generated
table(purrr::map_chr(landscapes, \(x) x$pattern))
```

### Selecting specific patterns

Specific patterns can be selected to create only a subset of patterns using the `patterns` argument. For example, to create landscapes with only labyrinth, spots, and clustered patterns:

```{r}
#| label: specific-patterns
#| fig-width: 8
#| fig-height: 6
#| fig-alt: "Grid showing landscapes with only labyrinth, spots, and clustered patterns"

# Generate only specific patterns
landscapes <- create_landscapes(
  n = 12,
  patterns = c("labyrinth", "spots", "clustered")
)

plot_landscapes(landscapes)
```

### Landscape size

To modify the size (number of pixels in x- and y-direction) of all generated landscapes, use the `width` and `height` arguments. By default, landscapes are created with a size of 100 x 100 pixels.

```{r}
#| label: custom-size
#| fig-width: 8
#| fig-height: 2
#| fig-alt: "Three non-square landscapes of 50 x 20 pixels"
non_square <- create_landscapes(
  n = 3,
  width = 50,
  height = 20
)
# Plot landscapes with custom size
plot_landscapes(non_square)
```

### Landscape rotation

By default, each rotatable pattern is rotated by a whole-degree angle sampled between 0 and 360. This exposes classifiers to similar landscapes in different orientations during training and helps prevent them from relying on orientation.

Set `rotation = 0` to disable rotation, or supply a length-2 range from which angles are sampled. This example uses angles between 45 and 90 degrees:

```{r}
#| label: rotation-example
#| message: false
#| fig-width: 8
#| fig-height: 3
#| fig-alt: "Three unrotated clustered, sharp, and banded landscapes followed by three clustered landscapes rotated between 45 and 90 degrees"

no_rotation <- create_landscapes(
  n = 3,
  patterns = c("clustered", "sharp", "bands"),
  rotation = 0
)

defined_angles <- create_landscapes(
  n = 3,
  patterns = "clustered",
  rotation = c(45, 90)
)

plot_landscapes(c(no_rotation, defined_angles))
```

Only "sharp", "diffuse", "fingers", "clustered" and "bands" are rotated. The remaining patterns ignore `rotation`, since a batch can mix rotatable and non-rotatable patterns.

### Pattern-specific parameters

The `params_list` argument sets parameters by pattern name using the same `pattern_*()` constructors. Unlike `create_landscape()`, a batch parameter can be either a length-2 range to sample the parameter from or a single value used for all landscapes of that pattern:

```{r}
#| label: custom-parameters
#| message: false
#| fig-width: 8
#| fig-height: 6
#| fig-alt: "Grid showing landscapes with many large spots of varying size"

bigger_spots <- create_landscapes(
  n = 12,
  patterns = "spots",
  params_list = list(
    spots = pattern_spots(
      n_spots = c(10, 20), # range for each landscape
      spot_radius = c(8, 12), # range for each landscape
      spot_radius_sd = 3, # single value for all landscapes
      regular_spots = FALSE # single value for all landscapes
    )
  )
)

plot_landscapes(bigger_spots)
```

A single value fixes a parameter for every landscape in the batch; a length-2 vector is a range. Above, `n_spots` and `spot_radius` vary between landscapes, while `spot_radius_sd` and `regular_spots` are the same throughout. Patterns left out of `params_list` fall back to their default sampling ranges, which are listed on each constructor's help page.

## Landscape objects

`create_landscape()` returns one `landscape` object, while `create_landscapes()` returns a list of them. Downstream package functions accept these objects directly.

Each landscape contains:

- `data`: landscape raster as a `SpatRaster`
- `pattern`: pattern label, such as `"spots"` or `"clustered"`
- `name`: user-defined name
- `params`: parameters used to generate the landscape

To convert your own matrices or rasters to this format, see [Import user-defined landscapes](importing-landscapes.html).

## Next steps

- [Calculate landscape metrics](landscape-metrics.html): Extract landscape features and identify informative metrics
- [Classify with landscape metrics](classify-metrics.html): Train classifiers on landscape metrics
- [Classify with pixels](classify-pixels.html): Train Keras classifiers on raster-cell values
