---
title: "Generate different landscapes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generate different landscapes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

This vignette shows how to generate artificial landscapes with different spatial patterns. These landscapes are designed for training and testing spatial pattern classifiers.

You can:

- Generate batches of training landscapes with `create_training_landscapes()`
- Create individual landscapes with custom parameters using `create_landscape()`

### Available patterns

There are 11 different spatial patterns available:

#### Control patterns

- **random**: Randomly distributed vegetation
- **bare**: No vegetation (control)
- **dense**: Complete vegetation cover (control)

```{r, echo=FALSE, fig.width=8, fig.height=3, fig.alt="Three landscapes showing control patterns: random, bare, and dense"}
random <- create_landscape("random", name = "Random")
bare <- create_landscape("bare", name = "Bare")
dense <- create_landscape("dense", name = "Dense")

plot_landscape_list(list(random, bare, dense), show_legend = FALSE)

```

#### Ecotone patterns

- **sharp**: Sharp treeline boundary
- **diffuse**: Gradual treeline transition
- **fingers**: Curvy finger-like extensions
- **clustered**: Vegetation clusters above treeline
- **bands**: Sinusoidal vegetation bands

```{r, echo=FALSE, fig.width=8, fig.height=5, fig.alt="Five landscapes showing ecotone patterns: sharp, diffuse, fingers, clustered, and bands"}
sharp <- create_landscape("sharp", name = "Sharp treeline")
diffuse <- create_landscape("diffuse", name = "Diffuse treeline")
fingers <- create_landscape("fingers", name = "Fingers")
clustered <- create_landscape("clustered", name = "Clustered")
bands <- create_landscape("bands", name = "Bands")
plot_landscape_list(
  list(sharp, diffuse, fingers, clustered, bands),
  show_legend = FALSE,
  ncol = 5
)
```

#### Patch patterns

- **spots**: Circular vegetation patches
- **gaps**: Circular vegetation gaps
- **labyrinth**: Maze-like vegetation patterns

```{r, echo=FALSE, fig.width=8, fig.height=3, fig.alt="Three landscapes showing patch patterns: spots, gaps, and labyrinth"}
spots <- create_landscape("spots", name = "Spots")
gaps <- create_landscape("gaps", name = "Gaps")
labyrinth <- create_landscape("labyrinth", name = "Labyrinth")
plot_landscape_list(list(spots, gaps, labyrinth), show_legend = FALSE)
```

For detailed parameter descriptions and examples of each pattern type, see `?create_landscape` and the individual pattern help pages (e.g., `?create_landscape_spots`, `?create_landscape_clustered`).

## Setup

```{r setup}
library(spatPatClassifyR)
# For reproducibility
set.seed(123456)
```

## Creating Multiple Training Landscapes

The quickest way to generate training data is with `create_training_landscapes()`.
It generates multiple landscapes balanced across all available patterns.

```{r, fig.width = 8, fig.height = 6, fig.alt = "Grid showing 20 randomly generated landscapes with different spatial patterns"}
# Create 20 landscapes balanced across all pattern types
landscapes <- create_training_landscapes(n = 20)

# Plot all landscapes
plot_landscape_list(landscapes)
```

### Selecting Specific Patterns

You can also select only specific patterns to include in the generated landscapes:

```{r, fig.width = 8, fig.height = 6, fig.alt = "Grid showing landscapes with only labyrinth, spots, and clustered patterns"}
# Generate only specific patterns
landscapes <- create_training_landscapes(
  n = 12,
  patterns = c("labyrinth", "spots", "clustered")
)

plot_landscape_list(landscapes)
```

### Control landscape parameters

You can modify default parameters for all generated landscapes of a specific pattern type by passing a list of parameter values to the `pattern_params` argument. For example, to create landscapes with more spots of larger size:

```{r, fig.width = 8, fig.height = 6, fig.alt = "Grid showing landscapes with many large spots"}
# Custom parameters for spot patterns
pattern_params <- list(
  spots = list(
    n_spots = 15,
    spot_radius = 10,
    spot_radius_sd = 3
  )
)
landscapes <- create_training_landscapes(
  n = 12,
  patterns = "spots",
  pattern_params = pattern_params
)
```

To check parameter names and default values for each pattern type, see the individual pattern help pages (e.g., `?create_landscape_spots`, `?create_landscape_clustered`).

## Creating Individual Landscapes

For more control over landscape parameters, create individual landscapes with `create_landscape()`.

### Simple Treeline Patterns

```{r, fig.width = 8, fig.height = 3, fig.alt = "Three landscapes showing sharp and diffuse treeline patterns with different positions"}
# Sharp treeline at different positions
sharp <- create_landscape("sharp", name = "Sharp default")
sharp_high <- create_landscape(
  "sharp",
  name = "High treeline",
  treeline_position = 0.7
)

# Diffuse treeline with varying steepness
diffuse <- create_landscape(
  "diffuse",
  name = "Steep transition",
  steepness = 0.1
)

plot_landscape_list(list(sharp, sharp_high, diffuse))
```

### Spot Patterns

```{r, fig.width = 8, fig.height = 5, fig.alt = "Four landscapes showing spot patterns with different numbers and sizes of vegetation patches"}
# Default spots
spots_default <- create_landscape("spots", name = "Default")

# More spots with size variation
spots_many <- create_landscape(
  "spots",
  name = "Many variable spots",
  n_spots = 15,
  spot_radius = 8,
  spot_radius_sd = 2
)

# Regular grid of spots
spots_regular <- create_landscape(
  "spots",
  name = "Regular grid",
  n_spots = 15,
  spot_radius = 20,
  spot_radius_sd = 0,
  regular_spots = TRUE
)

plot_landscape_list(list(spots_default, spots_many, spots_regular))
```

### Clustered Patterns

```{r, fig.width = 8, fig.height = 5, fig.alt = "Three landscapes showing clustered vegetation with different cluster sizes and numbers"}
# Few large clusters
clustered_large <- create_landscape(
  "clustered",
  name = "Few large clusters",
  n_clusters = 5,
  cluster_radius = 12,
  scatter_zone_prop = 0.5
)

# Many small clusters
clustered_small <- create_landscape(
  "clustered",
  name = "Many small clusters",
  n_clusters = 30,
  cluster_radius = 2,
  scatter_zone_prop = 0.6
)

# Elongated clusters
clustered_horizontal <- create_landscape(
  "clustered",
  name = "Horizontal clusters",
  n_clusters = 15,
  cluster_radius = 5,
  elongation_x = 3,
  elongation_y = 1,
  scatter_zone_prop = 0.4
)

plot_landscape_list(list(
  clustered_large,
  clustered_small,
  clustered_horizontal
))
```

### Labyrinth Patterns

```{r, fig.width = 8, fig.height = 3, fig.alt = "Two landscapes showing labyrinth-like vegetation patterns with different frequencies"}
# Default labyrinth
labyrinth_default <- create_landscape("labyrinth", name = "Default")

# Higher frequency with multiple octaves
labyrinth_complex <- create_landscape(
  "labyrinth",
  name = "Complex",
  frequency = 8,
  octaves = 3,
  band_fuzziness = 0.05
)

plot_landscape_list(list(labyrinth_default, labyrinth_complex))
```

## Modifying Landscape Metadata

Change the pattern label or name of existing landscapes:

```{r}
# Modify metadata
landscapes[[1]] <- set_landscape_pattern(landscapes[[1]], "custom_pattern")
landscapes[[1]] <- set_landscape_name(landscapes[[1]], "My custom landscape")

landscapes[[1]]
```

## Available Pattern Types

The package includes 11 spatial patterns:

- **random**: Randomly distributed vegetation
- **bare**: No vegetation (control)
- **dense**: Complete vegetation cover (control)
- **sharp**: Sharp treeline boundary
- **diffuse**: Gradual treeline transition
- **fingers**: Curvy finger-like extensions
- **clustered**: Vegetation clusters above treeline
- **bands**: Sinusoidal vegetation bands
- **spots**: Circular vegetation patches
- **gaps**: Circular vegetation gaps
- **labyrinth**: Maze-like vegetation patterns

For detailed parameter descriptions and examples of each pattern type, see `?create_landscape` and the individual pattern help pages (e.g., `?create_landscape_spots`, `?create_landscape_clustered`).

## Rotating Landscapes

All landscape generators support rotation:

```{r, fig.width = 8, fig.height = 3, fig.alt = "Two landscapes showing the same pattern rotated at different angles"}
# Same pattern, different rotations
diffuse_0 <- create_landscape("diffuse", name = "0 degrees")
diffuse_45 <- create_landscape("diffuse", name = "45 degrees", rotation = 45)

plot_landscape_list(list(diffuse_0, diffuse_45))
```

## Next Steps

- See `vignette("landscape-metrics")` for extracting landscape features
- See `vignette("basic-workflow")` for training classifiers with your landscapes

