---
title: "Matching training and application data"
vignette: >
  %\VignetteIndexEntry{matching-training-application}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

A classifier is trained on landscapes of a particular geometry and then applied to
new ones. When the application landscapes differ from the training data, the two
workflows behave differently, and some differences quietly degrade predictions.
This article explains what can differ, how each workflow responds, and how to
prepare application data so results stay trustworthy.

```{r}
#| label: setup
library(spatPatClassifyR)
```

## Four ways data can differ

"Different" is not one thing. Training and application landscapes can disagree on
four independent axes, and they matter to different degrees:

| Axis | Meaning | Example |
|---|---|---|
| **Extent** | size of the area, in cells | train 100 × 100, apply 200 × 200 |
| **Resolution (grain)** | ground size of one cell | train on 1 m cells, apply to 30 m imagery |
| **Aspect ratio** | shape of the rectangle | train 100 × 100, apply 200 × 80 |
| **Shape** | rectangular vs. irregular outline | apply to a clipped, non-rectangular map section |

The package generates landscapes as rectangular rasters, so an *irregular* study
area (a catchment, a park boundary) arrives as a rectangle with `NA` cells outside
its outline.

## How each workflow responds

The two workflows react to these axes in almost opposite ways, so the package
guards each accordingly.

| | metric-based | pixel-based |
|---|---|---|
| Extent / resolution | metrics computed as-is; **warns** on a substantial mismatch | rows and columns **resampled** onto the training grid; absolute grid size is discarded |
| Aspect ratio | reflected in the metrics; **warns** on a large change | anisotropic resize distorts the pattern; **warns** |
| Irregular shape (`NA`) | `NA` treated as outside the landscape; works | `NA` rejected before prediction; **error** |
| Training set itself | **warns** if training landscapes are not uniform | **requires** identical rows, columns, and layer counts (error otherwise) |

### Metric-based workflow

`calculate_metrics()` records each landscape's geometry, and `train_metric_model()`
stores a summary of the training geometry in the model. When you later call
`apply_metric_model()`, it compares the application landscapes against that summary
and warns if they differ substantially in extent, resolution, or aspect ratio:

```
! 3/6 application landscapes differ from the training extent by more than 25%.
i Scale-dependent metrics (area, edge, patch counts) change with extent, so
  predictions for these landscapes may be unreliable.
```

Whether a mismatch actually hurts depends on **which metrics the model uses**.
Area-, edge-, and count-based metrics (total area, patch density, edge density,
mean patch area) scale strongly with extent; normalised configuration metrics
(aggregation index, contagion, cohesion, Shannon diversity) are largely
scale-invariant. A model built on scale-invariant metrics tolerates size
differences well; one built on scale-dependent metrics does not.

::: {.callout-tip}
If you expect your application landscapes to differ in size from the training
landscapes, prefer scale-invariant metrics when selecting features with
`evaluate_metrics()`.
:::

The same logic applies to the *training* set: if your training landscapes are not
all the same size, `train_metric_model()` warns, because scale-dependent metrics
then conflate the pattern with the landscape's size.

### Pixel-based workflow

The CNN has a fixed input size, so `apply_pixel_model()` **resamples** every
application landscape onto the training grid. This lets the function accept
different row and column counts by discarding the absolute grid size. It does not
make the CNN scale-invariant, and predictions can still change when resampling
changes the apparent size of landscape structures. Two cautions remain:

- **Aspect ratio.** Resizing a non-square landscape onto a (square) training grid
  stretches it anisotropically, distorting the pattern. `apply_pixel_model()` warns
  when this happens.
- **`NA` cells.** They cannot be interpreted as a habitat category. The function
  raises an error before resizing or prediction rather than passing them to the
  network.

```
x Cannot classify landscapes that contain NA cells.
```

The training side is stricter still: the pixel workflow accepts one categorical
raster layer per landscape, and all training landscapes must have the same
numbers of rows and columns since their arrays are stacked into one tensor.
`train_pixel_model()` checks this and aborts with a clear message rather than
letting the stacking fail cryptically.

Habitat categories must be numeric whole-number codes. Training fixes one input
channel for each code, and application rasters must use the same codes. A trained
category may be absent from an application landscape, but a previously unseen
code cannot be interpreted by the fitted model.

## Preparing application data

### Keep the training set uniform

Generate training landscapes at a single width, height, and layer count.
`create_landscapes()` does this by default; if you assemble training data yourself,
make all three dimensions consistent.

### Different size, same shape

The pixel workflow resamples automatically, but check that the structures remain
meaningfully comparable after rescaling. The metric workflow is suitable when the
selected metrics are scale-invariant (see above).

### Different resolution (grain)

A model trained on synthetic 1 m landscapes is not directly comparable to 30 m
imagery: a "patch" means something different at each grain. Resample the
application data to the training grain, or retrain on data at the target grain,
before classifying.

### Irregular (non-rectangular) study areas

An irregular area arrives as a rectangle with `NA` outside its boundary. The metric
workflow handles this directly. The pixel workflow cannot (it errors on `NA`), so
reshape the data first. Two options:

1. **Crop** to the largest rectangle that fits inside the study area.
2. **Tile** the area into complete, training-sized windows and classify each,
   giving a spatial map of pattern types.

::: {.callout-warning}
Do **not** fill `NA` with `0` to make the raster complete. Zero is bare ground, so
this fabricates habitat that was never there and changes the very pattern you are
trying to classify.
:::

A simple crop to the largest centred square, so a later resize is not distorting:

```{r}
#| label: crop-to-square
#| fig-width: 8
#| fig-height: 4
# A non-square landscape (e.g. a wide map section)
wide <- create_landscape("sharp", width = 150, height = 90, name = "original")

# Crop to the largest centred square
raster <- wide$data
side <- min(terra::nrow(raster), terra::ncol(raster))
col_off <- floor((terra::ncol(raster) - side) / 2)
row_off <- floor((terra::nrow(raster) - side) / 2)
cropped <- terra::crop(
  raster,
  terra::ext(col_off, col_off + side, row_off, row_off + side)
)

square <- landscape(
  terra::as.matrix(cropped, wide = TRUE),
  name = "cropped",
  pattern = wide$pattern
)

plot_landscapes(list(wide, square), show_legend = FALSE)
```

The cropped landscape now matches the training aspect ratio, so
`apply_pixel_model()` resizes it without distortion.

## In short

- The metric workflow is sensitive to **extent and resolution** (via scale-dependent
  metrics) but handles **irregular shapes**; it warns on geometry mismatch.
- The pixel workflow accepts different row and column counts by **resampling** but
  does not preserve absolute scale. It needs **rectangular, `NA`-free,
  same-aspect** data to avoid errors or geometric distortion.
- Keep training landscapes uniform, choose scale-appropriate metrics, and crop or
  tile irregular areas. Never fill `NA` with `0`.
