---
title: "Matching training and application data"
vignette: >
  %\VignetteIndexEntry{matching-training-application}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

A classifier is trained on landscapes of a particular geometry and then applied to
new ones. When the application landscapes differ from the training data, the two
workflows behave differently, and some differences quietly degrade predictions.
This article explains what can differ, how each workflow responds, and how to
prepare application data so results stay trustworthy.

```{r}
#| label: setup
library(spatPatClassifyR)
```

## Four ways data can differ

"Different" is not one thing. Training and application landscapes can disagree on
four independent axes, and they matter to different degrees:

| Axis | Meaning | Example |
|---|---|---|
| **Extent** | size of the area, in cells | train 100 × 100, apply 200 × 200 |
| **Resolution (grain)** | ground size of one cell | train on 1 m cells, apply to 30 m imagery |
| **Aspect ratio** | shape of the rectangle | train 100 × 100, apply 200 × 80 |
| **Shape** | rectangular vs. irregular outline | apply to a clipped, non-rectangular map section |

The package generates landscapes as rectangular rasters, so an *irregular* study
area (a catchment, a park boundary) arrives as a rectangle with `NA` cells outside
its outline.

## How each workflow responds

The two workflows react to these axes in almost opposite ways, so the package
guards each accordingly.

| | metric-based | pixel-based |
|---|---|---|
| Extent / resolution | metrics computed as-is; **warns** on a substantial mismatch | landscape **resized** onto the training grid; robust |
| Aspect ratio | reflected in the metrics; **warns** on a large change | anisotropic resize distorts the pattern; **warns** |
| Irregular shape (`NA`) | `NA` treated as outside the landscape; works | `NA` reaches the CNN as `NaN`; **error** |
| Training set itself | **warns** if training landscapes are not uniform | **requires** identical training dimensions (error otherwise) |

### Metric-based workflow

`calculate_metrics()` records each landscape's geometry, and `train_metric_model()`
stores a summary of the training geometry in the model. When you later call
`apply_metric_model()`, it compares the application landscapes against that summary
and warns if they differ substantially in extent, resolution, or aspect ratio:

```
! 3/6 application landscapes differ from the training extent by more than 25%.
i Scale-dependent metrics (area, edge, patch counts) change with extent, so
  predictions for these landscapes may be unreliable.
```

Whether a mismatch actually hurts depends on **which metrics the model uses**.
Area-, edge-, and count-based metrics (total area, patch density, edge density,
mean patch area) scale strongly with extent; normalised configuration metrics
(aggregation index, contagion, cohesion, Shannon diversity) are largely
scale-invariant. A model built on scale-invariant metrics tolerates size
differences well; one built on scale-dependent metrics does not.

::: {.callout-tip}
If you expect your application landscapes to differ in size from the training
landscapes, prefer scale-invariant metrics when selecting features with
`evaluate_metrics()`.
:::

The same logic applies to the *training* set: if your training landscapes are not
all the same size, `train_metric_model()` warns, because scale-dependent metrics
then conflate the pattern with the landscape's size.

### Pixel-based workflow

The CNN has a fixed input size, so `apply_pixel_model()` **resizes** every
application landscape onto the training grid. This makes it robust to differences
in extent and resolution — it effectively discards absolute scale. Two cautions
remain:

- **Aspect ratio.** Resizing a non-square landscape onto a (square) training grid
  stretches it anisotropically, distorting the pattern. `apply_pixel_model()` warns
  when this happens.
- **`NA` cells.** They survive resizing and reach the network as `NaN`, which yields
  confident but meaningless predictions. Rather than fail silently,
  `apply_pixel_model()` raises an error.

```
x Cannot classify landscapes that contain NA cells.
```

The training side is stricter still: all training landscapes must have the same
cell dimensions, since their arrays are stacked into one tensor.
`train_pixel_model()` checks this and aborts with a clear message rather than
letting the stacking fail cryptically.

## Preparing application data

### Keep the training set uniform

Generate training landscapes at a single width and height. `create_landscapes()`
does this by default; if you assemble training data yourself, make the dimensions
consistent.

### Different size, same shape

No action needed. The pixel workflow resizes automatically, and the metric
workflow is fine as long as the selected metrics are scale-invariant (see above).

### Different resolution (grain)

A model trained on synthetic 1 m landscapes is not directly comparable to 30 m
imagery: a "patch" means something different at each grain. Resample the
application data to the training grain, or retrain on data at the target grain,
before classifying.

### Irregular (non-rectangular) study areas

An irregular area arrives as a rectangle with `NA` outside its boundary. The metric
workflow handles this directly. The pixel workflow cannot (it errors on `NA`), so
reshape the data first. Two options:

1. **Crop** to the largest rectangle that fits inside the study area.
2. **Tile** the area into complete, training-sized windows and classify each,
   giving a spatial map of pattern types.

::: {.callout-warning}
Do **not** fill `NA` with `0` to make the raster complete. Zero is bare ground, so
this fabricates habitat that was never there and changes the very pattern you are
trying to classify.
:::

A simple crop to the largest centred square, so a later resize is not distorting:

```{r}
#| label: crop-to-square
#| fig-width: 8
#| fig-height: 4
# A non-square landscape (e.g. a wide map section)
wide <- create_landscape("sharp", width = 150, height = 90, name = "original")

# Crop to the largest centred square
raster <- wide$data
side <- min(terra::nrow(raster), terra::ncol(raster))
col_off <- floor((terra::ncol(raster) - side) / 2)
row_off <- floor((terra::nrow(raster) - side) / 2)
cropped <- terra::crop(
  raster,
  terra::ext(col_off, col_off + side, row_off, row_off + side)
)

square <- landscape(
  terra::as.matrix(cropped, wide = TRUE),
  name = "cropped",
  pattern = wide$pattern
)

plot_landscapes(list(wide, square), show_legend = FALSE)
```

The cropped landscape now matches the training aspect ratio, so
`apply_pixel_model()` resizes it without distortion.

## In short

- The metric workflow is sensitive to **extent and resolution** (via scale-dependent
  metrics) but handles **irregular shapes**; it warns on geometry mismatch.
- The pixel workflow is robust to **extent and resolution** (it resizes) but needs
  **rectangular, `NA`-free, same-aspect** data; it errors or warns otherwise.
- Keep training landscapes uniform, choose scale-appropriate metrics, and crop or
  tile irregular areas — never fill `NA` with `0`.
```