---
title: "Pattern gallery"
vignette: >
  %\VignetteIndexEntry{pattern-gallery}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

An overview of the 11 spatial patterns available for artificial landscapes, organized into control, ecotone, and patch patterns. The examples show how selected parameters affect their appearance. To see all parameters for a specific pattern, open the corresponding help page, for example `?pattern_spots`.

See [Create artificial landscapes](landscape-generation.html) for instructions on generating individual landscapes and training batches.

```{r}
#| label: setup
library(patternscaper)
# Set seed for reproducible examples
set.seed(123456)
```

## Control patterns

The patterns **bare**, **random**, and **dense** place vegetation at random. They differ only in `veg_prob`, the probability that each cell is vegetated: `bare` defaults to 0.1, `random` to 0.5, and `dense` to 0.9. These patterns provide control landscapes without a specific spatial structure.

```{r}
#| label: control-patterns
#| fig-width: 8
#| fig-height: 3
#| fig-alt: "Three landscapes showing control patterns: bare, random, and dense"

bare <- create_landscape("bare", name = "bare (veg_prob 0.1)")
random <- create_landscape("random", name = "random (veg_prob 0.5)")
dense <- create_landscape("dense", name = "dense (veg_prob 0.9)")

list(bare, random, dense) |>
  plot_landscapes(
    titles = "name",
    show_legend = FALSE
  )
```

## Ecotone patterns

Ecotone patterns have a vegetated zone and a bare zone and a transition between them. In an unrotated landscape, the parameter `boundary_position` gives the relative position of this transition measured from the top. It applies to all five ecotone patterns.

### Sharp

An abrupt boundary, with `boundary_position` as its only parameter.

```{r}
#| label: pattern-sharp
#| fig-width: 8
#| fig-height: 3
#| fig-alt: "Two sharp-boundary landscapes: default and boundary position 0.3"

list(
  create_landscape("sharp", name = "default"),
  create_landscape(
    "sharp",
    name = "boundary_position = 0.3",
    params = pattern_sharp(boundary_position = 0.3)
  )
) |>
  plot_landscapes(
    titles = "name",
    show_legend = FALSE
  )
```

### Diffuse

A transition where vegetation probability decreases with distance below the boundary. `steepness` controls how quickly that probability drops: values near 0 produce a sparse, abrupt transition, while values near 1 produce a denser, more gradual transition.

```{r}
#| label: pattern-diffuse
#| fig-width: 8
#| fig-height: 3
#| fig-alt: "Three diffuse-boundary landscapes with increasingly dense transition zones"

list(
  create_landscape(
    "diffuse",
    name = "steepness = 0.1 (sparse)",
    params = pattern_diffuse(steepness = 0.1)
  ),
  create_landscape("diffuse", name = "default (0.5)"),
  create_landscape(
    "diffuse",
    name = "steepness = 1 (dense)",
    params = pattern_diffuse(steepness = 1)
  )
) |>
  plot_landscapes(
    titles = "name",
    show_legend = FALSE
  )
```

### Fingers

Curved, finger-like extensions of vegetation into the bare zone. The parameter `sine_length_mean` controls the average wavelength of the curves (larger values result in more widely spaced fingers). The parameter `sine_height_mean` controls the average finger amplitude (larger values result in longer fingers). The corresponding `*_sd` parameters control the variation between fingers.

```{r}
#| label: pattern-fingers
#| fig-width: 8
#| fig-height: 3
#| fig-alt: "Three finger landscapes: default, long pronounced bends, and short pronounced bends"

list(
  create_landscape("fingers", name = "default"),
  create_landscape(
    "fingers",
    name = "long, pronounced bends",
    params = pattern_fingers(
      sine_length_mean = 40,
      sine_height_mean = 12
    )
  ),
  create_landscape(
    "fingers",
    name = "short, pronounced bends",
    params = pattern_fingers(
      sine_length_mean = 10,
      sine_height_mean = 15,
      sine_height_sd = 3
    )
  )
) |>
  plot_landscapes(
    titles = "name",
    show_legend = FALSE
  )
```

### Clustered

Vegetation clusters scattered below the vegetation boundary in the bare zone. `n_clusters` and `cluster_radius` set cluster number and size, `cluster_zone` controls how far into the bare zone they can occur, and `elongation_x` and `elongation_y` stretch them along either axis.

```{r}
#| label: pattern-clustered
#| fig-width: 8
#| fig-height: 3
#| fig-alt: "Three clustered landscapes: default, few large clusters, and many small elongated clusters"

list(
  create_landscape("clustered", name = "default"),
  create_landscape(
    "clustered",
    name = "few, large",
    params = pattern_clustered(n_clusters = 5, cluster_radius = 10)
  ),
  create_landscape(
    "clustered",
    name = "many, small, elongated",
    params = pattern_clustered(
      n_clusters = 20,
      cluster_radius = 3,
      elongation_x = 3
    )
  )
) |>
  plot_landscapes(
    titles = "name",
    show_legend = FALSE
  )
```

### Bands

Sinusoidal vegetation bands parallel to the vegetation boundary. The parameters `band_thickness` and `band_spacing` control their geometry (how thick they are and how wide apart), `amplitude` sets the wave height, and `frequency` the amount of waves.

```{r}
#| label: pattern-bands
#| fig-width: 8
#| fig-height: 3
#| fig-alt: "Three banded landscapes: default, thick widely spaced bands, and bands with high, frequent waves"

list(
  create_landscape("bands", name = "default"),
  create_landscape(
    "bands",
    name = "thick, widely spaced",
    params = pattern_bands(band_thickness = 6, band_spacing = 20)
  ),
  create_landscape(
    "bands",
    name = "high, frequent waves",
    params = pattern_bands(amplitude = 15, frequency = 0.15)
  )
) |>
  plot_landscapes(
    titles = "name",
    show_legend = FALSE
  )
```

## Patch patterns

Patch patterns are self-organized patterns without a boundary between two zones.

### Spots

Circular vegetation patches on bare ground. `n_spots` and `spot_radius` set the count and size, `spot_radius_sd` varies the size between spots, `radius_noise_fraction` makes the spot edges rough, and `regular_spots = TRUE` places them on a grid instead of at random.

```{r}
#| label: pattern-spots
#| fig-width: 8
#| fig-height: 3
#| fig-alt: "Three spot landscapes: default, fewer larger variable spots, and a regular grid"

list(
  create_landscape("spots", name = "default"),
  create_landscape(
    "spots",
    name = "fewer, larger, variable",
    params = pattern_spots(
      n_spots = 3,
      spot_radius = 12,
      spot_radius_sd = 3,
      radius_noise_fraction = 0.2
    )
  ),
  create_landscape(
    "spots",
    name = "regular grid",
    params = pattern_spots(
      n_spots = 9,
      spot_radius = 8,
      regular_spots = TRUE
    )
  )
) |>
  plot_landscapes(
    titles = "name",
    show_legend = FALSE
  )
```

### Gaps

The inverse of spots: circular bare gaps in otherwise vegetated ground. It takes parameters analogous to those for spots, with names such as `n_gaps` and `gap_radius`.

```{r}
#| label: pattern-gaps
#| fig-width: 8
#| fig-height: 3
#| fig-alt: "Three gap landscapes: default, few large gaps, and many gaps with ragged edges"

list(
  create_landscape("gaps", name = "default"),
  create_landscape(
    "gaps",
    name = "few, large",
    params = pattern_gaps(n_gaps = 5, gap_radius = 15)
  ),
  create_landscape(
    "gaps",
    name = "ragged edges",
    params = pattern_gaps(
      n_gaps = 12,
      gap_radius = 8,
      radius_noise_fraction = 0.4
    )
  )
) |>
  plot_landscapes(
    titles = "name",
    show_legend = FALSE
  )
```

### Labyrinth

Maze-like bands of vegetation. The parameter `frequency` controls the scale of the maze, `octaves` how much fine detail is layered on top, and `veg_threshold` sets the overall vegetated area.

```{r}
#| label: pattern-labyrinth
#| fig-width: 8
#| fig-height: 3
#| fig-alt: "Three labyrinth landscapes: default, a finer maze, and a labyrinth with greater vegetation cover"

list(
  create_landscape("labyrinth", name = "default"),
  create_landscape(
    "labyrinth",
    name = "finer, more detailed",
    params = pattern_labyrinth(frequency = 6, octaves = 3)
  ),
  create_landscape(
    "labyrinth",
    name = "greater vegetation cover",
    params = pattern_labyrinth(veg_threshold = 0.35)
  )
) |>
  plot_landscapes(
    titles = "name",
    show_legend = FALSE
  )
```

## Next steps

- [Create artificial landscapes](landscape-generation.html): Select patterns and generate individual landscapes or training batches
