---
title: "Pattern gallery"
vignette: >
  %\VignetteIndexEntry{pattern-gallery}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

Compare the spatial patterns available for artificial landscapes. See [Create artificial landscapes](landscape-generation.html) for instructions on generating individual landscapes and training batches.

```{r}
#| label: setup
library(patternscaper)
# Set seed for reproducibility
set.seed(123456)
```

## The patterns

The package provides 11 spatial patterns. Each `pattern_*` help page documents the parameters you can set for the corresponding landscape.

### Control patterns

**random**, **bare**, and **dense** place vegetation at random. They differ only in `veg_prob`, the proportion of vegetated cells: `bare` defaults to 0.1, `random` to 0.5, and `dense` to 0.9. These patterns provide control landscapes without a specific spatial structure.

```{r}
#| label: control-patterns
#| fig-width: 8
#| fig-height: 3
#| fig-alt: "Three landscapes showing control patterns: bare, random, and dense"

bare <- create_landscape("bare", name = "bare (veg_prob 0.1)")
random <- create_landscape("random", name = "random (veg_prob 0.5)")
dense <- create_landscape("dense", name = "dense (veg_prob 0.9)")

plot_landscapes(
  list(bare, random, dense),
  titles = "name",
  show_legend = FALSE
)
```

### Ecotone patterns

Ecotone patterns have a vegetated zone and a bare zone separated by a transition. `boundary_position` places that transition in all five.

#### sharp

An abrupt boundary, with `boundary_position` as its only parameter.

```{r}
#| label: pattern-sharp
#| fig-width: 8
#| fig-height: 3
#| fig-alt: "Two sharp-boundary landscapes: default and a low boundary"

plot_landscapes(
  list(
    create_landscape("sharp", name = "default"),
    create_landscape(
      "sharp",
      name = "boundary_position = 0.3",
      params = pattern_sharp(boundary_position = 0.3)
    )
  ),
  titles = "name",
  show_legend = FALSE
)
```

#### diffuse

A gradual transition, where the chance of a cell being vegetated falls off with distance from the boundary. `steepness` runs from 0 to 1 and sets how wide that transition zone is: **higher** values spread it over more of the landscape.

```{r}
#| label: pattern-diffuse
#| fig-width: 8
#| fig-height: 3
#| fig-alt: "Three diffuse-boundary landscapes with increasingly wide transition zones"

plot_landscapes(
  list(
    create_landscape(
      "diffuse",
      name = "steepness = 0.1 (narrow)",
      params = pattern_diffuse(steepness = 0.1)
    ),
    create_landscape("diffuse", name = "default (0.5)"),
    create_landscape(
      "diffuse",
      name = "steepness = 1 (broad)",
      params = pattern_diffuse(steepness = 1)
    )
  ),
  titles = "name",
  show_legend = FALSE
)
```

#### fingers

Curving, finger-like extensions of vegetation into the bare zone. The parameter `sine_length_mean` controls the average wavelength of the curves (larger values result in longer, more widely spaced bends). The parameter `sine_height_mean` controls the average finger amplitude (larger values result in more pronounced bends). The corresponding `*_sd` parameters control the variation between fingers.

```{r}
#| label: pattern-fingers
#| fig-width: 8
#| fig-height: 3
#| fig-alt: "Three finger landscapes: default, long pronounced bends, and short pronounced bends"

plot_landscapes(
  list(
    create_landscape("fingers", name = "default"),
    create_landscape(
      "fingers",
      name = "long, pronounced bends",
      params = pattern_fingers(
        sine_length_mean = 40,
        sine_height_mean = 12
      )
    ),
    create_landscape(
      "fingers",
      name = "short, pronounced bends",
      params = pattern_fingers(
        sine_length_mean = 10,
        sine_height_mean = 15
      )
    )
  ),
  titles = "name",
  show_legend = FALSE
)
```

#### clustered

Vegetation clusters scattered above the boundary. `n_clusters` and `cluster_radius` set how many and how big, `cluster_zone` how far into the bare zone they reach, and `elongation_x` / `elongation_y` stretch them along one axis.

```{r}
#| label: pattern-clustered
#| fig-width: 8
#| fig-height: 3
#| fig-alt: "Three clustered landscapes: default, few large clusters, and many small elongated clusters"

plot_landscapes(
  list(
    create_landscape("clustered", name = "default"),
    create_landscape(
      "clustered",
      name = "few, large",
      params = pattern_clustered(n_clusters = 5, cluster_radius = 10)
    ),
    create_landscape(
      "clustered",
      name = "many, small, elongated",
      params = pattern_clustered(
        n_clusters = 20,
        cluster_radius = 3,
        elongation_x = 3
      )
    )
  ),
  titles = "name",
  show_legend = FALSE
)
```

#### bands

Sinusoidal vegetation bands running parallel to the boundary. `band_thickness` and `band_spacing` set their geometry, `amplitude` sets the height of their waves, and `frequency` sets how often they wave.

```{r}
#| label: pattern-bands
#| fig-width: 8
#| fig-height: 3
#| fig-alt: "Three banded landscapes: default, thick widely spaced bands, and bands with high, frequent waves"

plot_landscapes(
  list(
    create_landscape("bands", name = "default"),
    create_landscape(
      "bands",
      name = "thick, widely spaced",
      params = pattern_bands(band_thickness = 6, band_spacing = 20)
    ),
    create_landscape(
      "bands",
      name = "high, frequent waves",
      params = pattern_bands(amplitude = 15, frequency = 0.15)
    )
  ),
  titles = "name",
  show_legend = FALSE
)
```

### Patch patterns

Self-organized patterns, without a boundary between two zones.

#### spots

Circular vegetation patches on bare ground. `n_spots` and `spot_radius` set the count and size, `spot_radius_sd` varies the size between spots, `radius_noise_fraction` roughens their edges, and `regular_spots = TRUE` places them on a grid instead of at random.

```{r}
#| label: pattern-spots
#| fig-width: 8
#| fig-height: 3
#| fig-alt: "Three spot landscapes: default, fewer larger variable spots, and a regular grid"

plot_landscapes(
  list(
    create_landscape("spots", name = "default"),
    create_landscape(
      "spots",
      name = "fewer, larger, variable",
      params = pattern_spots(
        n_spots = 3,
        spot_radius = 12,
        spot_radius_sd = 3
      )
    ),
    create_landscape(
      "spots",
      name = "regular grid",
      params = pattern_spots(
        n_spots = 9,
        spot_radius = 8,
        regular_spots = TRUE
      )
    )
  ),
  titles = "name",
  show_legend = FALSE
)
```

#### gaps

The inverse of spots: circular bare gaps in otherwise vegetated ground. It takes parameters analogous to those for spots, with names such as `n_gaps` and `gap_radius`.

```{r}
#| label: pattern-gaps
#| fig-width: 8
#| fig-height: 3
#| fig-alt: "Three gap landscapes: default, few large gaps, and many gaps with ragged edges"

plot_landscapes(
  list(
    create_landscape("gaps", name = "default"),
    create_landscape(
      "gaps",
      name = "few, large",
      params = pattern_gaps(n_gaps = 5, gap_radius = 15)
    ),
    create_landscape(
      "gaps",
      name = "ragged edges",
      params = pattern_gaps(
        n_gaps = 12,
        gap_radius = 8,
        radius_noise_fraction = 0.4
      )
    )
  ),
  titles = "name",
  show_legend = FALSE
)
```

#### labyrinth

Maze-like bands of vegetation. `frequency` sets the scale of the maze and `octaves` how much fine detail is layered on top; `veg_threshold` shifts the overall vegetated proportion.

```{r}
#| label: pattern-labyrinth
#| fig-width: 8
#| fig-height: 3
#| fig-alt: "Three labyrinth landscapes: default, a finer maze, and a labyrinth with greater vegetation cover"

plot_landscapes(
  list(
    create_landscape("labyrinth", name = "default"),
    create_landscape(
      "labyrinth",
      name = "finer, more detailed",
      params = pattern_labyrinth(frequency = 6, octaves = 3)
    ),
    create_landscape(
      "labyrinth",
      name = "greater vegetation cover",
      params = pattern_labyrinth(veg_threshold = 0.35)
    )
  ),
  titles = "name",
  show_legend = FALSE
)
```

