---
title: "pattern-showcase"
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

```{r}
#| label: setup
library(spatPatClassifyR)
```


# Introduction
Brief overview of landscape patterns and their ecological relevance.

# Pattern Overview

The package provides 13 distinct landscape patterns divided into three categories.
Each pattern can be customized with specific parameters, which we explore in detail below.

## Boundary Patterns

Represent transitions between two classes (e.g., forest and non-forest) along boundaries:

```{r boundary-patterns, fig.width=10, fig.height=6}
boundary_patterns <- c(
  "sharp",
  "diffuse",
  "fingers",
  "bands",
  "clustered"
)

boundary_showcase <- boundary_patterns |>
  purrr::map(\(pattern) create_landscape(pattern, name = pattern))

plot_landscape_list(boundary_showcase, title = "name", show_legend = FALSE)
```

- `sharp`: Abrupt boundary
- `diffuse`: Gradual transition with scattered elements
- `fingers`: Irregular, finger-like projections
- `bands`: Horizontal bands with sinusoidal variation
- `clustered`: Grouped patches near a boundary

## Spatial Structure Patterns

Complex spatial arrangements and textures without clear boundaries:

```{r spatial-patterns, fig.width=10, fig.height=6}
spatial_patterns <- c("labyrinth", "spots", "gaps")

spatial_showcase <- spatial_patterns |>
  purrr::map(\(pattern) create_landscape(pattern, name = pattern))

plot_landscape_list(spatial_showcase, title = "name", show_legend = FALSE)
```

- `labyrinth`: Maze-like perlin noise patterns
- `spots`: Circular patches of one class
- `gaps`: Inverse of spots (circular patches of opposite class)

## Random Patterns

Uniform random distribution with controllable density:

```{r random-patterns, fig.width=10, fig.height=3}
# Show random with different densities
random_patterns <- c("bare", "random", "dense")

random_showcase <- random_patterns |>
  purrr::map(\(pattern) create_landscape(pattern, name = pattern))

plot_landscape_list(random_showcase, title = "name", show_legend = FALSE)
```

- `random`: Uniform random distribution 
- `bare`: Low density of class 1 (10%)
- `dense`: High density of class 1 (90%)


# Boundary Patterns

## Sharp Boundary
[description + manual parameter examples]

## Diffuse Boundary
[description + manual parameter examples]

## Fingers
[description + manual parameter examples]

# Spatial Structure Patterns
Complex spatial arrangements and textures.

## Random (+ Bare/Dense)
[description + manual parameter
[description + manual parameter examples]

## Labyrinth
[description + manual parameter examples]

## Spots
[description + manual parameter examples]

## Gaps
[description + manual parameter examples]

# Generating Training Landscapes
How to use `create_training_landscapes()` for multiple patterns with variation.
